---
title: "energyRt — interpolation & solve settings comparison"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
params:
  model_file: "tmp/espt.RData"   # RData holding the model object (relative to project root)
  model_object: "mod"            # name of the model object inside model_file
  horizon_period: 2024           # newHorizon(period = ...) applied to every build
  solve: false                   # FALSE = interpolation comparison only; TRUE = also solve
  solvers: ["julia_highs_barrier", "pyomo_cbc"]  # solver_options names (lists scale; GLPK only for small models)
  solve_settings: null           # NULL = solve every setting; or a subset of setting names
  glpk_path: "c:/Users/admin/.conda/envs/multimod/Library/bin"      # set if GLPK isn't on PATH
  julia_path: "c:/Users/admin/AppData/Local/Microsoft/WindowsApps/" # set for julia_* solvers
  python_path: "c:/Users/admin/.conda/envs/pypsa-eur"               # set for pyomo_* solvers
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      collapse = TRUE)
# Load the in-development package (knit with knit_root_dir = package root, so
# "." is the package). Falls back to the installed package.
if (requireNamespace("devtools", quietly = TRUE) && file.exists("DESCRIPTION")) {
  suppressMessages(devtools::load_all(".", quiet = TRUE))
} else {
  suppressMessages(library(energyRt))
}
suppressMessages(library(knitr))

# Load the model object (`params$model_object`) from `params$model_file`.
stopifnot(file.exists(params$model_file))
.e <- new.env(); load(params$model_file, envir = .e)
mod <- get(params$model_object, envir = .e)

# Apply the requested horizon (period) to every build.
hz <- newHorizon(period = params$horizon_period)

# Point the package at the solver toolchains needed by the requested solvers.
if (isTRUE(params$solve)) {
  sv <- params$solvers
  if (any(grepl("glpk",  sv)) && nzchar(params$glpk_path  %||% ""))
    try(set_glpk_path(params$glpk_path),   silent = TRUE)
  if (any(grepl("julia", sv)) && nzchar(params$julia_path %||% ""))
    try(set_julia_path(params$julia_path), silent = TRUE)
  if (any(grepl("pyomo", sv)) && nzchar(params$python_path %||% ""))
    try(set_python_path(params$python_path), silent = TRUE)
}
```

This report interpolates a model under several storage settings and compares the
resulting size. When `params$solve = TRUE` it then solves each build with the
requested solver(s) and checks that the objective is invariant to the storage
settings.

```{r config, echo = FALSE}
# The interpolation settings grid compared below. Edit to taste.
settings <- list(
  sparse_prune         = list(fold = FALSE, sparse = TRUE, prune = TRUE),
  fold_sparse_prune    = list(fold = TRUE,  sparse = TRUE, prune = TRUE),
  foldall_sparse_prune = list(
    fold = c("region", "slice", "year", "comm", "tech", "stg", "trade"),
    sparse = TRUE, prune = TRUE)
)

cat("model       :", params$model_object, "from", params$model_file, "\n")
cat("horizon     : newHorizon(period =", params$horizon_period, ")\n")
cat("settings    :", paste(names(settings), collapse = ", "), "\n")
cat("solve       :", params$solve, "\n")
if (isTRUE(params$solve)) cat("solvers     :", paste(params$solvers, collapse = ", "), "\n")
```

# Interpolation comparison

```{r run}
if (isTRUE(params$solve)) {
  # Interpolate + solve in one pass (keeps scenarios), so we don't interpolate
  # twice. The interpolation-size table is `res$interp`.
  solve_grid <- if (is.null(params$solve_settings)) settings else
    settings[params$solve_settings]
  solver_list <- lapply(params$solvers, function(n) solver_options[[n]])
  res <- compare_solve_settings(mod, settings = solve_grid,
                                solvers = solver_list, horizon = hz)
  interp_tbl <- res$interp
  top_tbl    <- res$top
} else {
  res <- compare_interp_settings(mod, settings = settings, horizon = hz)
  interp_tbl <- res$summary
  top_tbl    <- res$top
}
```

Size of each interpolated build (in memory):

```{r interp-table, echo = FALSE}
kable(interp_tbl, format.args = list(big.mark = ","),
      caption = "Per-setting build size and time")
```

Largest parameters by row count, per setting (blank = folded to a wildcard or
not in the top list):

```{r top-table, echo = FALSE}
kable(utils::head(top_tbl, 15), format.args = list(big.mark = ","),
      caption = "Top parameters (rows) per setting")
```

# Solve comparison

```{r solve-table, echo = FALSE, eval = isTRUE(params$solve)}
kable(res$summary, format.args = list(big.mark = ","),
      caption = "Objective, time and status per setting x solver")
```

```{r solve-invariance, echo = FALSE, eval = isTRUE(params$solve)}
# The objective must NOT depend on storage settings: per solver it should match
# across fold / sparse / prune. (Printed by the comparison object.)
ok <- res$summary[res$summary$ok & !is.na(res$summary$objective), , drop = FALSE]
for (sl in unique(ok$solver)) {
  v <- ok$objective[ok$solver == sl]
  same <- length(v) < 2 ||
    diff(range(v)) <= 1e-6 * max(1, abs(stats::median(v)))
  cat(sprintf("%-12s objective: %s  [%s]\n", sl,
              paste(format(v, digits = 8), collapse = "  "),
              if (same) "invariant across settings" else "DIFFERS — investigate"))
}
```

```{r no-solve-note, echo = FALSE, eval = !isTRUE(params$solve)}
cat("Solving is disabled (params$solve = FALSE).\n",
    "Set `solve: true` (and `solvers`) in the YAML params to run the solve",
    "comparison.\n")
```
