---
title: "Introduction to energyRt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{energyRt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.retina = TRUE
)
```

```{r setup, include = FALSE}
library(energyRt)
set_scenarios_path(file.path(tempdir(), "intro"))
```

# The idea

An energy system is a network of **flows**: primary resources are extracted or
harvested, converted by technologies, moved, stored, and finally consumed as
useful energy. Planning such a system — what to build, when, and how to run
it — is an optimization problem: meet demand in every place and moment at the
lowest total cost, subject to physics and policy.

**energyRt** is a macro-language for stating that problem in R. You describe
the *system* — fuels, technologies, resources, demands — as R objects, and the
package compiles them into a full capacity-expansion and dispatch optimization
model: it derives the sets, writes the equations, runs a solver, and returns
the solution as tidy tables. The modeling layer (hundreds of index-mapping and
bookkeeping decisions) is generated, so your code stays at the level of the
energy system itself.

# One model, four backends

The energyRt optimization model — around one hundred predefined equations,
extendable with `newConstraint()` — is implemented in four mathematical
programming languages:

* **GLPK / MathProg** — open source; bundled with Rtools on Windows, so it
  works out of the box;
* **Julia / JuMP** — open source, with fast solvers like HiGHS;
* **Python / Pyomo** — open source, CBC/HiGHS and others;
* **GAMS** — the commercial standard in energy-systems work.

The *same* model object solves on any backend with consistent results — start
on GLPK with zero setup, switch to a faster or institutional solver later
without touching your model. The full mathematical formulation is documented
in the [model equations PDF](https://energyrt.org/articles/energyRt_model_equations.pdf);
the [solver backends article](https://energyrt.org/articles/backends.html)
covers configuration.

# The bricks

Models are assembled from a small set of object types:

| Brick | Constructor | Role |
|---|---|---|
| commodity | `newCommodity()` | an energy carrier or accounting flow (GAS, ELC, CO2) |
| technology | `newTechnology()` | a conversion process — the richest object |
| supply | `newSupply()` | a source: fuel at a price, or a free resource |
| demand | `newDemand()` | a sink: final consumption to be met |
| storage | `newStorage()` | shifts a commodity across time slices |
| weather | `newWeather()` | capacity factors and other exogenous profiles |
| repository | `newRepository()` | a bag of bricks — the parts library |
| model | `newModel()` | repository + regions + calendar + horizon |
| scenario | `interpolate_model()` / `solve_scenario()` | a model instance, solved |

A technology converts inputs to outputs through the chain
`input → use → activity → output`; costs, capacity, availability, emissions
and secondary flows all attach to it. The
[model bricks article](https://energyrt.org/articles/model-bricks.html)
walks through every object in depth.

# A five-minute model

One fuel, one power plant, one demand — a complete, solvable system:

```{r hello-build}
GAS <- newCommodity("GAS", timeframe = "ANNUAL")
ELC <- newCommodity("ELC", timeframe = "ANNUAL")

SUP_GAS <- newSupply("SUP_GAS", commodity = "GAS",
                     availability = data.frame(cost = 6.0))   # fuel price, MEUR/PJ

EGAS <- newTechnology("EGAS",
  input   = list(comm = "GAS"), output = list(comm = "ELC"),
  ceff    = data.frame(comm = "GAS", cinp2use = 0.55),        # 55% efficient
  invcost = list(invcost = 900),                              # MEUR/GW
  fixom   = 25, cap2act = 31.536, olife = 25L)

DEM_ELC <- newDemand("DEM_ELC", commodity = "ELC",
                     dem = data.frame(dem = 50))               # 50 PJ a year

draw(EGAS)
```

Bind the bricks to space and time, and solve on GLPK:

```{r hello-solve, results = "hide"}
mod <- newModel("HELLO",
  data    = newRepository("parts", GAS, ELC, SUP_GAS, EGAS, DEM_ELC),
  region  = "R1", discount = 0.05,
  horizon = newHorizon(period = 2025:2040, intervals = c(1, 5, 10),
                       mid_is_end = TRUE))

scen <- solve_scenario(interpolate_model(mod, name = "BASE"),
                       solver = solver_options$glpk,
                       echo = FALSE)    # echo = FALSE silences the solver log
```

Results come back as tidy tables:

```{r hello-results}
getData(scen, "vObjective", merge = TRUE)   # total discounted system cost, MEUR
getData(scen, "vTechCap",   merge = TRUE)   # capacity the model built, GW
```

The model built `r round(getData(scen, "vTechCap", merge = TRUE)$value[1], 2)`
GW of gas capacity — exactly enough to deliver 50 PJ a year — and the solution
reports every flow, cost, and capacity variable at the same level of detail.

# How it stands out

**One language end to end.** Data preparation, model formulation, solving, and
analysis all happen in R: your inputs arrive via the tidyverse, and results
return as tibbles ready for `dplyr` and `ggplot2`. There is no export-import
seam between the model and your analysis.

**Solver independence.** Institutional modeling ecosystems typically bind you
to one platform. energyRt's four equivalent backends mean a class can run on
GLPK, a research group on JuMP, and an agency on GAMS — same model, same
results.

**Built for scale.** A sparse interpolation engine materializes only the
set-tuples a scenario needs, and scenarios are stored in Arrow-backed folders
(`save_scenario()` / `load_scenario()`) that load lazily — larger-than-memory
results remain workable.

**Analysis built in.** `levcost()` prices a technology *a-priori* (a textbook
LCOE from a unit-demand mini-model) or *ex-post* from a solved scenario;
`report()` renders a technology or process datasheet; `draw()` diagrams any
technology; `autoplot()` covers calendars, commodities, weather, costs and
more.

**A complete teaching model.** The UTOPIA electricity model ships with the
package — both as step-by-step vignettes and as the ready `utopia_modules`
data kit with scenario levers (CO~2~ cap, carbon tax, renewable share, nuclear
moratorium) — so the path from "hello world" to policy analysis is paved.

energyRt sits alongside established energy-modeling frameworks (the
TIMES/MESSAGE/OSeMOSYS family) as an R-native alternative: the same class of
capacity-expansion models, formulated and analyzed without leaving R.

# Where next

1. [Model bricks](https://energyrt.org/articles/model-bricks.html) — every
   object type in depth, including fuel blends, auxiliary flows, and user
   constraints.
2. `vignette("utopia-build")` — *UTOPIA I*: build a multi-region electricity
   model brick by brick.
3. `vignette("utopia-use")` — *UTOPIA II*: solve it and run policy scenarios.
4. [Solver backends](https://energyrt.org/articles/backends.html) — GLPK,
   JuMP, Pyomo, GAMS, and NEOS.
5. [Workflow](https://energyrt.org/articles/workflow.html) — extracting,
   editing, saving and comparing scenarios.
```{r cleanup, include = FALSE}
unlink(file.path(tempdir(), "intro"), recursive = TRUE)
```
