---
title: "UTOPIA II: running scenarios"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{utopia-use}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
# eval = FALSE: this vignette is frozen (code shown, not executed on build) --
# it drives full GLPK solves and is validated out-of-band. Flip to TRUE to run.
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, fig.retina = TRUE
)
# defined here (this chunk always runs) so `eval = have_sf` chunk options below
# resolve while the vignette is frozen; set TRUE when un-freezing to build maps.
have_sf <- FALSE
```

# Overview

This vignette uses the packaged [`utopia_modules`] kit (built in *UTOPIA I:
building the model modules*) to assemble an energy-system model and run
scenarios with **GLPK** (bundled, no external solver needed).

```{r setup}
library(energyRt)
library(dplyr)
library(ggplot2)
have_sf <- requireNamespace("sf", quietly = TRUE)   # region maps need the sf package
set_scenarios_path(file.path(tempdir(), "utopia"))
set_default_solver(solver_options$glpk)             # set once, used by every solve
```

Two calls run every scenario in this vignette: `interpolate_model()` expands
the model over regions, years and timeslices; `solve_scenario()` writes the solver
files, runs the default solver (set once above) and reads the solution.
`echo = FALSE` keeps the solver log out of the page — drop it to watch GLPK
work.

# Selecting a UTOPIA structure

`utopia_modules$electricity` offers ready region layouts (`reg1`, `reg3`,
`reg7`); each is a kit with a base repository `$repo` and scenario levers.
Calendars, horizons and maps sit alongside. (For a custom layout or calendar,
build the blocks yourself following *UTOPIA I: building the model*.)

```{r select}
names(utopia_modules$electricity)          # available layouts
um  <- utopia_modules$electricity$R3     # the 3-region base case
cal <- utopia_modules$calendars$utopia_s4h24
hor <- utopia_modules$horizons$base        # milestones 2020/2030/2040/2050
```

# Assembling and solving the base model

```{r base}
mod <- newModel("UTOPIA", data = um$repo, calendar = cal,
                region = um$regions, horizon = hor, discount = 0.05)
scen_BASE <- interpolate_model(mod, name = "BASE") |>
  solve_scenario(echo = FALSE)
getData(scen_BASE, "vObjective", merge = TRUE)$value    # total system cost, MEUR
```

The `reg3` kit wires its regions with bi-directional transmission links
(`TBD_ELC_*`). `plot_trade_map()` draws that network over a map layout — pass any
of `utopia_modules$maps` (here `honeycomb`); the 7-region kit shows six links:

```{r trade-map, eval = have_sf, fig.height = 4}
plot_trade_map(um$repo, map = utopia_modules$maps$honeycomb)
```

Electricity generation by technology:

```{r gen}
getData(scen_BASE, "vTechOut", comm = "ELC", merge = TRUE) |>
  group_by(year, tech) |> summarise(PJ = sum(value), .groups = "drop") |>
  ggplot(aes(factor(year), PJ, fill = tech)) + geom_col() +
  labs(x = "year", title = "Generation by technology (BASE)") + theme_bw()
```

The dispatch over a representative summer day -- generation stacked by
technology, storage charging below zero, demand as the line
(`autoplot(scenario)` reads the solution via `getMix()`; note that `getData()`
aggregates sub-annual values to ANNUAL by default -- pass
`timeframe = "highest"` for timeslice-level data):

```{r storage}
autoplot(scen_BASE, "generation", timeslice = "^SUM_")
```

# Scenarios: layering the levers

Each scenario adds a pre-built lever to the base model. The pattern is always
`interpolate_model(mod, name, LEVER)` then `solve_scenario()`:

```{r scenarios}
scen_CO2CAP <- interpolate_model(mod, "CO2CAP", um$CO2_CAP) |>       # emissions cap
  solve_scenario(echo = FALSE)
scen_CTAX <- interpolate_model(mod, "CARBONTAX", um$CT_CO2) |>       # carbon tax
  solve_scenario(echo = FALSE)
scen_RES <- interpolate_model(mod, "RES_SHARE", um$RES_SHARE) |>     # renewable target
  solve_scenario(echo = FALSE)
scen_NUC <- interpolate_model(mod, "NO_NEW_NUC", um$NO_NEW_NUC) |>   # no new nuclear
  solve_scenario(echo = FALSE)
scen_ERET <- interpolate_model(mod, "EARLY_RET", um$EARLY_RET) |>    # early fossil retirement
  solve_scenario(echo = FALSE)
```

The fifth lever, `EARLY_RET`, puts a declining per-region ceiling on installed
coal and gas capacity -- the plants carry `optimizeRetirement = TRUE`, so the
model chooses *which* units to close first while the constraint sets the pace.
The ceiling bites hardest on gas (the flexible backstop for wind and solar),
pushing investment into nuclear, biomass and storage instead.

Compare the scenarios with `compare_scenarios()`: the overview carries every
run's objective (with differences against the baseline), the value tables
flag which solved variables moved, and `autoplot()` draws the comparison
charts. The full comparative document is one `report(cmp)` call away.

```{r compare-co2}
scl <- list(BASE = scen_BASE, CO2CAP = scen_CO2CAP,
            CARBONTAX = scen_CTAX, RES_SHARE = scen_RES,
            EARLY_RET = scen_ERET)
cmp <- compare_scenarios(scl)
cmp
autoplot(cmp, "emissions")
```

Total system cost (objective) by scenario, straight from the overview:

```{r compare-cost}
with(cmp$overview, setNames(round(objective), scenario))
```

(The same charts can always be hand-rolled from the tidy layer --
`getData(scl, ...)` and `getMix(scl, ...)` accept named scenario lists and
add a `scenario` column.)

A carbon **tax** and a CO2 **cap** both cut emissions but through different
mechanisms (price vs quantity); a **renewable target** raises cost and can even
raise CO2 relative to a carbon price -- a useful contrast for policy discussion.

# Levelized cost of a process

`levcost()` on a **solved scenario** returns the *ex-post* levelized cost of a
process: the discounted sum of its own costs (annualised investment `vTechEac`,
fixed and variable O&M, and attributed fuel) divided by its output in that
solution. Because it reads the actual results -- not a mini-model -- it is exact
at the model's time resolution, with no calendar caveat.

```{r levcost-base}
lc <- levcost(scen_BASE, name = "ENUC")      # the existing nuclear fleet
lc$levcost_npv                               # NPV levelized cost, MEUR/PJ
autoplot(lc, type = "components")            # eac / fixom / varom / supply, by year
```

`ENUC` runs on base-year capacity, so its **capital is already sunk** -- the
annualised-investment term `eac` is ~0 and the realized cost is **fixed O&M plus
fuel**; a *newly built* plant would add its `eac` here. The cost per PJ also
depends on how hard a plant is run: more output spreads the fixed costs, lowering
the levelized cost. Priced across scenarios, `ENUC`'s LCOE tracks how much the
existing fleet is dispatched as the policy mix changes:

```{r levcost-compare}
sapply(scl, function(s) {
  lc <- suppressMessages(levcost(s, name = "ENUC"))
  if (inherits(lc, "levcost")) round(as.numeric(lc$levcost_npv), 2) else NA_real_
})
```

(`levcost(scen, name)` attributes fuel from `vTechInp`, so it works for the
single-fuel technologies here; a plant with a *grouped* fuel input would need
its group activity instead.) A full technology datasheet with the ex-post cost
embedded is one call away (`report()` renders HTML/PDF; needs pandoc):

```{r levcost-report, eval = FALSE}
report(scen_BASE, name = "ENUC", format = "html")

# the whole-scenario results report: run provenance (scenario_runs()),
# solution checks, generation/capacity charts, and a role-driven cost
# breakdown; `run =` reports any recorded run, not only the active one
report(scen_BASE)
```

# Other structures

The same code runs on a different layout -- e.g. the single-region `reg1` or the
7-region `reg7` (heavier). Just swap the kit:

```{r other, eval = FALSE}
um7 <- utopia_modules$electricity$R7
mod7 <- newModel("UTOPIA7", data = um7$repo, calendar = cal,
                 region = um7$regions, horizon = hor, discount = 0.05)
scen7 <- interpolate_model(mod7, "BASE7") |>
  solve_scenario(echo = FALSE)
```

# Next steps

Extend UTOPIA with inter-regional trade (using the `utopia_modules$maps`),
technologies introduced via scenarios (CCS, hydrogen, CHP), and the honeycomb
multi-region layout. See the workshop exercises for a step-by-step build.
