---
title: "Plotting energyRt objects"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 2.8,
  fig.retina = 2,
  dpi = 110,
  # The plots require ggplot2 (a Suggests dependency); skip evaluation if absent.
  eval = requireNamespace("ggplot2", quietly = TRUE)
)
```

`energyRt` ships `autoplot()` methods (and, where natural, matching `plot()`
methods) for most of its building-block classes. They turn an object into a
`ggplot`, so the result can be themed and extended with `+` layers like any other
ggplot.

| Class | What the plot shows |
|-------|---------------------|
| `calendar` | the sub-annual time structure (stacked timeframe rows) |
| `horizon` | the planning intervals / milestone years |
| `commodity` | emission factors (`@emis`) — one or several commodities |
| `supply`, `demand`, `import`, `export` | level parameters over years (given data + interpolation) |
| `weather` | a sub-annual capacity/availability factor (heatmap, line or area) |

```{r setup}
library(energyRt)
library(ggplot2)   # provides the autoplot() generic
```

# Calendars

The example `calendars` dataset holds a few ready-made `calendar` objects.

```{r}
data("calendars", package = "energyRt")
names(calendars)
```

`autoplot()` (or the equivalent `plot()`) draws each timeframe as a row of
rectangles sized by each slice's share of the year:

```{r cal-basic, fig.height = 2.2}
autoplot(calendars$season_dn)
```

## Coloring

For `fill = "order"` (the default) the `color_pattern` controls the gradient.
`"within"` (default) colours each level over **its own** slices — so an hourly
row shows a full `h00`→`h23` cycle repeating every day, while `"global"` colours
by absolute chronology across the whole year.

```{r cal-color, fig.height = 2.6}
autoplot(calendars$d365_h24)                          # within (default)
autoplot(calendars$d365_h24, color_pattern = "global")
```

Other fill metrics are the slice `"share"` and `"weight"`:

```{r cal-share, fig.height = 2.2}
autoplot(calendars$season_dn, fill = "share")
```

## Labels

Each cell can be labelled by its individual level name (`"name"`, the default,
e.g. `h00`…`h23`) or by the full slice path (`"slice"`, e.g. `d001_h00`). Label
text auto-contrasts with the fill; override it with `label_color`.

```{r cal-labels, fig.height = 2.6}
# zoom (see below) so labels are legible
autoplot(calendars$d365_h24,
         show_leafs = list(YDAY = "d001", HOUR = 1:24),
         label_by = "name")
```

## Zooming with `show_leafs`

`show_leafs` selects which slices to draw. Pass a named list to filter per
timeframe level (character = slice names, numeric = positions among that level's
slices), or an unnamed vector to filter the finest level directly.

```{r cal-zoom, fig.height = 2.4}
# day 100, hours 5-10
autoplot(calendars$d365_h24, show_leafs = list(YDAY = "d100", HOUR = 5:10))

# the first 48 leaf slices (first two days)
autoplot(calendars$d365_h24, show_leafs = 1:48)
```

## Subset view

A reduced calendar can be shown against a full reference calendar: pass the full
one as `reference`. The slices present in the subset are filled; the rest are
left empty.

```{r cal-subset, fig.height = 2.4}
autoplot(calendars$d365_h24_subset_1day_per_month,
         reference = calendars$d365_h24)
```

## Styling

`border` outlines the rectangles (off by default so dense rows read as a smooth
gradient) and `palette` picks the viridis option; and because the result is a
ggplot you can keep customizing it:

```{r cal-style, fig.height = 2.2}
autoplot(calendars$season_dn, palette = "magma", border = "grey40") +
  labs(title = "Four seasons, day/night") +
  theme_minimal()
```

# Heatmaps

`autoplot()` lays every slice out in one row per timeframe. When a value is
indexed by a *two-dimensional* calendar — day-of-year × hour, month × hour — a
**heatmap** reads more naturally: `plot_heatmap()` puts the finest timeframe on
`y`, the next on `x`, and any coarser levels into facets. The layout is taken
from the calendar (pass it as `calendar =`), or guessed from the slice names with
`tsl_guess_format()`.

Here is a synthetic hourly load profile on the `d365_h24` calendar (a daily cycle
plus a seasonal swing). The diurnal band and the winter/summer contrast are
immediately visible:

```{r hm-day, fig.height = 2.6}
cal  <- calendars$d365_h24
tt   <- cal@timetable
prof <- data.frame(
  slice = tt$slice,
  load  = 50 + 30 * sin(2 * pi * (tsl2hour(tt$slice) - 6) / 24) +
                15 * cos(2 * pi *  tsl2yday(tt$slice)      / 365))

plot_heatmap(prof, calendar = cal, value = "load")   # x = YDAY, y = HOUR
```

Facet a day-of-year format by `"month"` to get one day-of-month × hour panel per
month:

```{r hm-month, fig.height = 4.5}
plot_heatmap(prof, value = "load", facet = "month")  # 12 monthly panels
```

`palette` selects the viridis option and `name` labels the colour bar, exactly as
for `autoplot()`.

## From model objects

Any slice-indexed model data lays out the same way. Here **electricity demand**
and a **solar capacity factor** from the packaged UTOPIA kit, on its season × hour
calendar (`utopia_s4h24`): pull the object with `getObject()`, take one region
(and, for demand, one year), and pass the `slice` + value columns to
`plot_heatmap()`. The `name` argument carries the value's unit onto the colour
bar.

```{r hm-demand, fig.height = 2.6}
repo <- utopia_modules$electricity$reg3$repo
wcal <- calendars$utopia_s4h24

DEM <- getObject(repo, name = "DEM_ELC", drop = TRUE)
dem <- subset(as.data.frame(DEM@dem), region == "R1" & year == 2050)
plot_heatmap(dem[, c("slice", "dem")], calendar = wcal, value = "dem", name = "PJ")
```

```{r hm-weather, fig.height = 2.6}
WSOL <- getObject(repo, name = "WSOL", drop = TRUE)
wsol <- subset(as.data.frame(WSOL@weather), region == "R1")
plot_heatmap(wsol[, c("slice", "wval")], calendar = wcal, value = "wval",
             name = "capacity factor")
```

The weather heatmap is exactly what `autoplot()` draws for a `weather` object
directly — with line and area variants too (see *Weather* below).

# Horizons

`horizon` objects draw the modelled intervals over milestone years:

```{r hor, fig.height = 2.4}
data("horizons", package = "energyRt")
autoplot(horizons$Y2020_2060_by_5)
```

# Commodities

For a `commodity`, `autoplot()` shows its emission factors (`@emis`). Passing
several commodities compares their emission intensities side by side.

```{r comm}
COA <- newCommodity("COA", desc = "Coal",
                    emis = data.frame(comm = "CO2", unit = "kt/GWh", emis = 0.33))
OIL <- newCommodity("OIL", emis = data.frame(comm = "CO2", unit = "kt/GWh", emis = 0.25))
GAS <- newCommodity("GAS", emis = data.frame(comm = "CO2", unit = "kt/GWh", emis = 0.18))

autoplot(COA)                 # a single commodity
autoplot(COA, OIL, GAS)       # comparison
```

# Supply, demand, import, export

These process objects are plotted by year: the **points** are the given data and
the **lines** are the interpolation (both extracted with `getData()`). Level
parameters are faceted by their base name so bounds and prices/costs keep
separate y-scales.

```{r sup, fig.height = 3}
SUP_COA <- newSupply(
  name = "SUP_COA", commodity = "COA", unit = "PJ", region = "R1",
  availability = data.frame(
    region = "R1", slice = "ANNUAL",
    year   = c(2020, 2040, 2020, 2020),
    ava.up = c(100,  200,  NA,   NA),
    ava.lo = c(NA,   NA,   10,   NA),
    cost   = c(NA,   NA,   NA,   30)))
autoplot(SUP_COA)
```

A single given value (or none) is drawn as a flat dashed line showing the
interpolation direction. Pass `years` to interpolate over a specific horizon:

```{r sup-years, fig.height = 3}
autoplot(SUP_COA, years = 2020:2050)
```

Demand, import and export follow the same pattern:

```{r dem, fig.height = 2.6}
DEM_ELC <- newDemand(
  name = "DEM_ELC", commodity = "ELC", unit = "GWh",
  dem = data.frame(region = "R1", slice = "ANNUAL",
                   year = c(2020, 2030, 2050), dem = c(100, 150, 300)))
autoplot(DEM_ELC)
```

```{r imp, fig.height = 3}
IMP_GAS <- newImport(
  name = "IMP_GAS", commodity = "GAS",
  imp = data.frame(region = "R1", slice = "ANNUAL",
                   year = c(2020, 2050), imp.up = c(50, 80), price = c(5, 9)))
autoplot(IMP_GAS)
```

# Weather

`weather` objects hold a sub-annual factor — a capacity / availability factor —
per region and slice. `autoplot()` offers three views via `type =`: a calendar
**heatmap** (the default), and diurnal **line** / **area** charts. Because the
slice layout (here season × hour) can't be inferred from the slice names alone,
pass the model's `calendar`; the value's unit (`@unit`, or `"capacity factor"`
when unset) labels the value axis — the fill legend for the heatmap, the y-axis
for line/area.

```{r wea, fig.height = 3}
WSOL <- getObject(utopia_modules$electricity$reg3$repo, name = "WSOL", drop = TRUE)
wcal <- calendars$utopia_s4h24

autoplot(WSOL, calendar = wcal)                    # heatmap (default), faceted by region
```

The **line** view reads the diurnal shape directly — one line per season, the
capacity factor on the y-axis:

```{r wea-line, fig.height = 3}
autoplot(WSOL, type = "line", calendar = wcal)
autoplot(WSOL, type = "area", calendar = wcal)     # or filled areas
```

# Notes

* Every method returns a `ggplot`, so `+ theme_*()`, `+ labs()`, `+ scale_*()`
  etc. all work.
* `plot()` is available for `calendar` and `horizon` and produces the same figure
  as `autoplot()`.
* `autoplot()` also works on the result of `levcost()` (levelized cost) — see
  `?levcost` — but that requires solving a small model, so it is not shown here.
* See `?autoplot.calendar`, `?autoplot.commodity`, `?autoplot.supply`, and
  `?plot_weather` for the full list of arguments.
