---
title: "Solver backends and solver_options"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

energyRt formulates the optimization model **once** and then writes it out for
one of several mathematical-programming **backends**, runs the solver, and reads
the solution back. You pick the backend and the underlying solver with a single
**`solver_options` preset**; the same model and scenario code works with all of
them, which makes cross-checking results between backends easy.

```{r setup}
library(energyRt)
```

# The backends at a glance

| Backend | `lang` | Solvers | License | Notes |
|---|---|---|---|---|
| **GLPK / MathProg** | `GLPK` | GLPK | open-source | Ships as the `glpsol` executable; easiest to install, slow on large models. |
| **Python / Pyomo** | `PYOMO` | CBC, GLPK, CPLEX | open-source (CBC/GLPK) | Convenient with a conda environment. |
| **Julia / JuMP** | `JuMP` | HiGHS, Cbc, GLPK, CPLEX | open-source (HiGHS/Cbc/GLPK) | HiGHS barrier is fast; recommended for large models. |
| **Julia / cuOpt** (GPU) | `JuMP` | NVIDIA cuOpt | open-source | **Linux only.** First-order (PDLP) needs no factorization, so it fits models whose barrier does not; add crossover for duals. |
| **GAMS** | `GAMS` | CPLEX, CBC | proprietary | Needs a GAMS license; supports GDX I/O and fine solver tuning. |
| **NEOS** (remote) | `PYOMO` / `GAMS` | CPLEX, CBC, … | free service | Solve on the [NEOS server](https://neos-server.org) — no local commercial solver needed. |
| **multimod_cloud** (cloud GPU) | `PYOMO` (as MPS) | HiGHS / cuOpt on a cloud GPU | service | **Experimental.** Writes the problem as MPS and solves it remotely, for models too large for local RAM. |

See the **Installation and Settings** article for how to install each runtime
and the library layer (`en_check_dependencies()`, `en_install_deps()`).

# Choosing a backend

`solver_options` is a named list of ready-made presets — each one pins a backend,
a solver, the data-exchange format, and any solver-specific tuning:

```{r}
names(solver_options)
solver_options$glpk
solver_options$julia_highs_barrier
```

Use a preset in one of two ways:

```{r}
# 1. Set it as the session default, then solve normally
set_default_solver(solver_options$julia_highs)
get_default_solver()                       # inspect the current default
scen <- solve_model(model)                   # uses the default

# 2. Pass it explicitly to a single solve
scen <- solve_model(model,  solver = solver_options$glpk)
scen <- solve_scenario(scen,  solver = solver_options$gams_gdx_cplex)
```

`set_default_solver()` lasts for the session. To make a default stick across
sessions, save it with `en_config_write()`; `en_config_show()` reports the
current default together with the solver paths it depends on.

`solve_model()` interpolates a `model` and solves it; `solve_scenario()` solves an
already-interpolated scenario. With `solver = NULL` the scenario's own settings
or `get_default_solver()` are used.

# The backends in detail

## GLPK / MathProg

The GNU Linear Programming Kit. energyRt writes the model in GMPL/MathProg and
calls the `glpsol` executable — the lowest-friction option and a good default for
small and medium models.

```{r}
solve_model(model, solver = solver_options$glpk)
```

## Python / Pyomo

energyRt writes a Pyomo `ConcreteModel` and solves it with the configured
solver. CBC is the open-source default; CPLEX and GLPK are also available.
Model data and the solution are exchanged as Arrow files, which needs
`pyarrow` in the Python environment (`en_install_python_deps()` installs it);
`solver_options$pyomo_cbc_sqlite` selects the older SQLite exchange instead.

```{r}
solve_model(model, solver = solver_options$pyomo_cbc)
solve_model(model, solver = solver_options$pyomo_cplex_barrier)   # CPLEX barrier
```

## Julia / JuMP

energyRt writes a JuMP model and solves it in Julia. **HiGHS** (open-source) is
fast and the recommended choice for large models; `*_barrier`, `*_simplex` and
`*_parallel` presets select the HiGHS algorithm. Cbc, GLPK and CPLEX are also
available. Model data and the solution are exchanged as Arrow files, read and
written by `Arrow.jl` (`en_install_julia_pkgs()` installs it);
`solver_options$julia_highs_rdata` selects the older RData exchange instead.

```{r}
solve_model(model, solver = solver_options$julia_highs)           # default (auto)
solve_model(model, solver = solver_options$julia_highs_barrier)   # interior point
solve_model(model, solver = solver_options$julia_highs_parallel)  # parallel simplex
```

## NVIDIA cuOpt (GPU)

**Linux only.** cuOpt solves the same JuMP model on an NVIDIA GPU through
[`cuOpt.jl`](https://github.com/jump-dev/cuOpt.jl). It ships no Windows build,
so on Windows the Julia process must run inside WSL. Install cuOpt itself
first (`pip install --extra-index-url=https://pypi.nvidia.com cuopt-cu12`),
put `libcuopt.so` on `LD_LIBRARY_PATH`, then add the wrapper with
`en_install_julia_pkgs("cuOpt")`. `cuOpt.jl` supports one cuOpt release at a
time, so the installed `cuopt-cu12` must match it.

```{r}
solve_model(model, solver = solver_options$julia_cuopt_pdlp)      # first-order, GPU
solve_model(model, solver = solver_options$julia_cuopt_crossover) # + basic solution
solve_model(model, solver = solver_options$julia_cuopt_concurrent)# race all three
```

`method` selects the algorithm: `0` races PDLP and barrier on the GPU against
dual simplex on the CPU and returns whichever finishes first, `1` is PDLP alone.
Which preset to use follows from how the solution is consumed:

* **`julia_cuopt_pdlp`** -- PDLP never factorizes, so GPU memory grows linearly
  with non-zeros. This is the preset for models whose barrier factorization does
  not fit in GPU memory. The solution carries a duality gap of about `1e-4`,
  which is fine for capacities and dispatch but not for prices.
* **`julia_cuopt_crossover`** -- adds a crossover to a basic solution, which is
  what duals (and therefore shadow prices) require. Costs a little extra time
  and recovers the exact optimum.
* **`julia_cuopt_concurrent`** -- fastest on small and medium models, but the
  barrier leg dominates GPU memory and the winning algorithm can differ between
  runs, so the same model may not return byte-identical results twice.

## GAMS

If you have a GAMS license, energyRt can write and run a GAMS model. The
`gams_gdx_*` presets exchange data as GDX (fast, native); the `gams_csv_*`
presets use text/CSV. The `*_barrier` and `*_parallel` presets inject a tuned
`cplex.opt` (LP method, threads, crossover, …).

```{r}
solve_model(model, solver = solver_options$gams_gdx_cplex)
solve_model(model, solver = solver_options$gams_gdx_cplex_barrier)
```

## NEOS (remote solve)

The [NEOS Server](https://neos-server.org) runs commercial solvers (CPLEX, …)
for free over the internet. The model is **built locally** but the **solve is
dispatched to NEOS**, so you can use CPLEX without a local license. Two families
are provided — via Pyomo (`neos_pyomo_*`) and via GAMS (`neos_gams_*`, sent as
inlined text so no local GAMS install is needed).

NEOS requires an email address, set once via the `NEOS_EMAIL` environment
variable or the helper (this is the one energyRt setting whose environment
variable is *not* prefixed `ENERGYRT_`, because the Pyomo subprocess reads
`NEOS_EMAIL` directly):

```{r}
set_neos_email("you@example.org")
solve_model(model, solver = solver_options$neos_gams_cplex)
solve_model(model, solver = solver_options$neos_pyomo_cplex_barrier)
```

The low-level NEOS interface (submit/poll/fetch a job, list solvers, …) is also
exported — see `?neos` and the `neos_*` functions.

## multimod_cloud (cloud GPU)

**Experimental — in testing.** Where NEOS dispatches the *model text*, this
route writes the problem as an **MPS file locally** and sends that to a GPU
solver in the cloud, decoding the solution back into the scenario. It exists
for models whose build or solve does not fit a laptop: the MPS is produced by
the Pyomo formulation, so nothing new has to be written for the model itself.

```{r}
solve_model(model, solver = solver_options$multimod_cloud)
# or, to recover duals from a first-order solve:
solve_model(model, solver = solver_options$multimod_cloud_crossover)
```

Being experimental, treat results as provisional and cross-check a smaller
variant against a local backend before relying on them. Access is via a token
(`get_hf_token()`), and the preset carries the machine flavour and a wall-clock
timeout, so a run that exceeds it returns nothing rather than partial results.

# All preset `solver_options`

The presets shipped with the package (generated from the installed data):

```{r presets, eval = TRUE, echo = FALSE}
so <- energyRt::solver_options
field <- function(x, nm, default = "—") if (is.null(x[[nm]])) default else x[[nm]]
lang_label <- c(GLPK = "GLPK/MathProg", PYOMO = "Python/Pyomo",
                JuMP = "Julia/JuMP", GAMS = "GAMS")
tab <- do.call(rbind, lapply(so, function(x) data.frame(
  preset   = x$name,
  backend  = unname(lang_label[field(x, "lang", NA)]),
  solver   = field(x, "solver"),
  # NOT keyed on `backend`: that marks the R-side GAMS/NEOS route only, while
  # `neos_pyomo_*` dispatches inside Python via `inc_solver`. The name is the
  # reliable marker for both.
  remote   = if (grepl("^neos_", x$name)) "NEOS"
             else if (identical(field(x, "backend", ""), "multimod_cloud"))
               "cloud GPU" else "",
  exchange = field(x, "export_format", "default"),
  stringsAsFactors = FALSE)))
knitr::kable(tab, row.names = FALSE,
             col.names = c("preset", "backend", "solver", "remote", "data exchange"))
```

A `default` in the last column means the preset names no format and follows the
`exchange_format` option — Arrow IPC unless you change it.

# Data exchange and storage

Two separate option families, easy to confuse because both once shared one
option:

* **exchange** — the files handed to a solver for one solve, in
  `runs/[<variant>/]<solve>/input|output/`. Written once, read once, deleted with the
  run. `get_exchange_format()` (`"feather"`), `get_exchange_compression()`
  (`"lz4"` — the codec is cheap because nothing here is kept).
* **storage** — the tables that outlive the session: a scenario's parameter and
  variable stores, and the model / repository / dataset stores.
  `get_storage_format()` (`"feather"`), `get_storage_compression()`
  (`"zstd"`, level 15).

```{r}
set_exchange_format("parquet")     # Pyomo only; Arrow.jl cannot read parquet
set_exchange_compression("uncompressed")
set_storage_format("parquet")      # smaller on disk; feather reads faster
```

A solver preset's `export_format` / `import_format` overrides the exchange
option for that solve. A store records the format it was written in and keeps
it when rewritten, so changing the option never mixes codecs inside an existing
store — it applies to what you write next.

# Anatomy of a preset

Each preset is a plain list, so you can copy one and tweak it. Common fields:

| Field | Meaning |
|---|---|
| `name` | preset name (informational) |
| `lang` | backend: `"GLPK"`, `"PYOMO"`, `"JuMP"`, `"GAMS"` |
| `solver` | underlying solver (e.g. `"HiGHS"`, `"cbc"`, `"CPLEX"`) |
| `export_format` / `import_format` | data exchange in / out: `feather` (Arrow, the default), `parquet` (Pyomo only), `SQLite` / `RData` (legacy input), `GDX`, CSV/text. Unset on a JuMP/Pyomo preset means "follow `get_exchange_format()`"; the two directions are independent |
| `backend` | `"neos"` for a remote NEOS solve |
| `neos_solver`, `neos_category` | remote solver and NEOS problem category |
| `inc3`, `inc4`, … | backend-specific code blocks injected into the model template (e.g. a `cplex.opt` for LP-method tuning) |

```{r}
# start from a preset and customise
my_solver <- solver_options$julia_highs
my_solver$name <- "julia_highs_custom"
# ... adjust an inc* block to set HiGHS attributes ...
solve_model(model, solver = my_solver)
```

# See also

* **Installation and Settings** — installing the runtimes and library layer.
* **Model bricks** — assembling commodities, processes and structures into a model.
* `?solve_model`, `?solve_scenario`, `?set_default_solver`, `?set_neos_email`.
* `?energyRt-options` for every setting, `?en_config_show` for what is in effect.
