---
title: "Solver backends and solver_options"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

energyRt formulates the optimization model **once** and then writes it out for
one of several mathematical-programming **backends**, runs the solver, and reads
the solution back. You pick the backend and the underlying solver with a single
**`solver_options` preset**; the same model and scenario code works with all of
them, which makes cross-checking results between backends easy.

```{r setup}
library(energyRt)
```

# The backends at a glance

| Backend | `lang` | Solvers | License | Notes |
|---|---|---|---|---|
| **GLPK / MathProg** | `GLPK` | GLPK | open-source | Ships as the `glpsol` executable; easiest to install, slow on large models. |
| **Python / Pyomo** | `PYOMO` | CBC, GLPK, CPLEX | open-source (CBC/GLPK) | Convenient with a conda environment. |
| **Julia / JuMP** | `JuMP` | HiGHS, Cbc, GLPK, CPLEX | open-source (HiGHS/Cbc/GLPK) | HiGHS barrier is fast; recommended for large models. |
| **GAMS** | `GAMS` | CPLEX, CBC | proprietary | Needs a GAMS license; supports GDX I/O and fine solver tuning. |
| **NEOS** (remote) | `PYOMO` / `GAMS` | CPLEX, CBC, … | free service | Solve on the [NEOS server](https://neos-server.org) — no local commercial solver needed. |

See the **Installation and Settings** article for how to install each runtime
and the library layer (`en_check_dependencies()`, `en_install_deps()`).

# Choosing a backend

`solver_options` is a named list of ready-made presets — each one pins a backend,
a solver, the data-exchange format, and any solver-specific tuning:

```{r}
names(solver_options)
solver_options$glpk
solver_options$julia_highs_barrier
```

Use a preset in one of two ways:

```{r}
# 1. Set it as the session default, then solve normally
set_default_solver(solver_options$julia_highs)
get_default_solver()                       # inspect the current default
scen <- solve_mod(model)                   # uses the default

# 2. Pass it explicitly to a single solve
scen <- solve_mod(model,  solver = solver_options$glpk)
scen <- solve_scen(scen,  solver = solver_options$gams_gdx_cplex)
```

`solve_mod()` interpolates a `model` and solves it; `solve_scen()` solves an
already-interpolated scenario. With `solver = NULL` the scenario's own settings
or `get_default_solver()` are used.

# The backends in detail

## GLPK / MathProg

The GNU Linear Programming Kit. energyRt writes the model in GMPL/MathProg and
calls the `glpsol` executable — the lowest-friction option and a good default for
small and medium models.

```{r}
solve_mod(model, solver = solver_options$glpk)
```

## Python / Pyomo

energyRt writes a Pyomo `ConcreteModel` (data exchanged as SQLite by default,
or Arrow/feather with the `*_arrow` presets) and solves it with the configured
solver. CBC is the open-source default; CPLEX and GLPK are also available.

```{r}
solve_mod(model, solver = solver_options$pyomo_cbc)
solve_mod(model, solver = solver_options$pyomo_cplex_barrier)   # CPLEX barrier
```

## Julia / JuMP

energyRt writes a JuMP model and solves it in Julia. **HiGHS** (open-source) is
fast and the recommended choice for large models; `*_barrier`, `*_simplex` and
`*_parallel` presets select the HiGHS algorithm. Cbc, GLPK and CPLEX are also
available.

```{r}
solve_mod(model, solver = solver_options$julia_highs)           # default (auto)
solve_mod(model, solver = solver_options$julia_highs_barrier)   # interior point
solve_mod(model, solver = solver_options$julia_highs_parallel)  # parallel simplex
```

## GAMS

If you have a GAMS license, energyRt can write and run a GAMS model. The
`gams_gdx_*` presets exchange data as GDX (fast, native); the `gams_csv_*`
presets use text/CSV. The `*_barrier` and `*_parallel` presets inject a tuned
`cplex.opt` (LP method, threads, crossover, …).

```{r}
solve_mod(model, solver = solver_options$gams_gdx_cplex)
solve_mod(model, solver = solver_options$gams_gdx_cplex_barrier)
```

## NEOS (remote solve)

The [NEOS Server](https://neos-server.org) runs commercial solvers (CPLEX, …)
for free over the internet. The model is **built locally** but the **solve is
dispatched to NEOS**, so you can use CPLEX without a local license. Two families
are provided — via Pyomo (`neos_pyomo_*`) and via GAMS (`neos_gams_*`, sent as
inlined text so no local GAMS install is needed).

NEOS requires an email address, set once via the `NEOS_EMAIL` environment
variable or the helper:

```{r}
set_neos_email("you@example.org")
solve_mod(model, solver = solver_options$neos_gams_cplex)
solve_mod(model, solver = solver_options$neos_pyomo_cplex_barrier)
```

The low-level NEOS interface (submit/poll/fetch a job, list solvers, …) is also
exported — see `?neos` and the `neos_*` functions.

# All preset `solver_options`

The presets shipped with the package (generated from the installed data):

```{r presets, eval = TRUE, echo = FALSE}
so <- energyRt::solver_options
field <- function(x, nm, default = "—") if (is.null(x[[nm]])) default else x[[nm]]
lang_label <- c(GLPK = "GLPK/MathProg", PYOMO = "Python/Pyomo",
                JuMP = "Julia/JuMP", GAMS = "GAMS")
tab <- do.call(rbind, lapply(so, function(x) data.frame(
  preset   = x$name,
  backend  = unname(lang_label[field(x, "lang", NA)]),
  solver   = field(x, "solver"),
  remote   = if (identical(field(x, "backend", ""), "neos")) "NEOS" else "",
  exchange = field(x, "export_format", "default"),
  stringsAsFactors = FALSE)))
knitr::kable(tab, row.names = FALSE,
             col.names = c("preset", "backend", "solver", "remote", "data exchange"))
```

# Anatomy of a preset

Each preset is a plain list, so you can copy one and tweak it. Common fields:

| Field | Meaning |
|---|---|
| `name` | preset name (informational) |
| `lang` | backend: `"GLPK"`, `"PYOMO"`, `"JuMP"`, `"GAMS"` |
| `solver` | underlying solver (e.g. `"HiGHS"`, `"cbc"`, `"CPLEX"`) |
| `export_format` / `import_format` | data exchange: `SQLite`, `feather` (Arrow), `GDX`, CSV/text |
| `backend` | `"neos"` for a remote NEOS solve |
| `neos_solver`, `neos_category` | remote solver and NEOS problem category |
| `inc3`, `inc4`, … | backend-specific code blocks injected into the model template (e.g. a `cplex.opt` for LP-method tuning) |

```{r}
# start from a preset and customise
my_solver <- solver_options$julia_highs
my_solver$name <- "julia_highs_custom"
# ... adjust an inc* block to set HiGHS attributes ...
solve_mod(model, solver = my_solver)
```

# See also

* **Installation and Settings** — installing the runtimes and library layer.
* **Model bricks** — assembling commodities, processes and structures into a model.
* `?solve_mod`, `?solve_scen`, `?set_default_solver`, `?set_neos_email`.
