---
title: "Installation and Settings"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

energyRt formulates an optimization model once and hands it to one of several
mathematical-programming **backends**. You only need *one* of them to get
started; you can add more later and cross-check results between them.

| Backend | Software to install | License | Notes |
|---|---|---|---|
| **GLPK** | `glpsol` executable | open-source | Easiest to install; slow on very large models. |
| **Julia / JuMP** | Julia + `JuMP`, `HiGHS` | open-source | Fast (HiGHS barrier); recommended for large models. |
| **Python / Pyomo** | Python + `pyomo` + a solver (CBC) | open-source | Convenient if you already use conda. |
| **GAMS** | GAMS distribution | proprietary | Needs a license; also enables GDX I/O. |

The package ships helpers to **detect** what you have and **auto-install the
library layer** (R/Julia/Python packages). The *system software* itself (Julia,
Python, conda, GAMS) is installed by you, with guidance below.

## 1. Install R and energyRt

Install R (≥ 4.3) — platform installers:
[Windows](https://cran.r-project.org/bin/windows/base/),
[macOS](https://cran.r-project.org/bin/macosx/),
[Ubuntu/Debian](https://cran.r-project.org/bin/linux/ubuntu/). We recommend
[RStudio](https://posit.co/download/rstudio-desktop/) as the IDE.

The quickest install is one line — it sets up `pak`, reports any Linux system
libraries to install, installs all dependencies, then energyRt itself:

```{r}
source("https://raw.githubusercontent.com/optimal2050/energyRt/master/inst/install.R")
install_energyRt()
```

Then verify your setup in a fresh session with `library(energyRt); en_setup()`.

Prefer to install directly? energyRt's CRAN dependencies are pulled in
automatically by `pak` (no local sibling packages are required):

```{r}
install.packages("pak")
pak::pkg_install("optimal2050/energyRt")

# or, with remotes:
# install.packages("remotes")
# remotes::install_github("optimal2050/energyRt")
```

Installing energyRt does not install a solver: you still need at least one
backend (below). The GAMS GDX bridge (`gdxtools`) is pulled later by
`en_install_deps()` if you use the GAMS backend.

**Install fails on Linux?** If the install stops with
`dependencies '...' are not available for package 'energyRt'`, one CRAN import
(often one that needs a system library) failed and aborted the rest. Install the
imports one by one so the failure is isolated and reported:

```{r}
install.packages("pak")

# energyRt's direct CRAN imports (base packages are omitted):
deps <- c(
  "generics", "data.table", "DBI", "RSQLite", "tibble", "tidyr", "dplyr",
  "rlang", "stringr", "lubridate", "purrr", "arrow", "progressr", "tictoc",
  "cli", "zoo", "registry", "options", "glue", "plyr",
  # suggested — plots and reports (optional but recommended, see section 7):
  "ggplot2", "patchwork", "knitr", "rmarkdown", "tinytex", "sf"
)

failed <- character(0)
for (pkg in deps) {
  ok <- tryCatch(
    {
      pak::pkg_install(pkg, ask = FALSE)
      message("  [ok]   ", pkg)
      TRUE
    },
    error = function(e) {
      message("  [FAIL] ", pkg, ": ", conditionMessage(e))
      FALSE
    }
  )
  if (!ok) failed <- c(failed, pkg)
}

if (length(failed) == 0) {
  message("All dependencies installed — installing energyRt itself:")
  pak::pkg_install("optimal2050/energyRt")
} else {
  message("Failed packages: ", paste(failed, collapse = ", "))
}
```

On Linux, a package that fails to compile is missing a system library. Ask
`pak` which apt packages are required, install them, then re-run the loop:

```{r}
pak::pkg_sysreqs("optimal2050/energyRt")   # lists the apt packages
# e.g. sudo apt install libcurl4-openssl-dev libssl-dev libxml2-dev
```

## 2. Check what you already have

```{r}
library(energyRt)
en_check_dependencies()   # solver backends: GLPK / Julia / Python / GAMS / GDX
en_check_packages()       # R packages, optional extras, LaTeX engine, MuseScore
```

`en_check_dependencies()` prints a status table — for each backend: installed?,
version, path, and a hint for what to do next — and tells you whether at least
one solver is ready. Each backend also has its own detector, e.g.
`en_check_julia()`, `en_check_python()`, `en_check_pyomo()`, `en_check_glpk()`,
`en_check_gams()`. `en_check_packages()` complements it by checking energyRt's R
dependencies, optional packages (plots/reports), and external tools (a LaTeX
engine, MuseScore). `en_setup()` runs both.

## 3. Install the system software (one or more backends)

Pick at least one. These are the only manual steps; commands are grouped by
operating system.

- **GLPK** — the `glpsol` executable.
  - *Windows*: **already included** — `glpsol.exe` ships with **Rtools 4.5**
    (which you already have, since Rtools is needed to build energyRt from
    source), at `C:/rtools45/x86_64-w64-mingw32.static.posix/bin`. No separate
    install.
  - *macOS*: `brew install glpk`.
  - *Ubuntu/Debian*: `sudo apt install glpk-utils`.
  - If `glpsol` is not detected, set the path —
    `set_glpk_path("C:/rtools45/x86_64-w64-mingw32.static.posix/bin")`
    (macOS/Ubuntu package managers put it on the `PATH` automatically).
- **Julia** — install via [juliaup](https://github.com/JuliaLang/juliaup).
  - *Windows*: `winget install julia -s msstore` (or the official installer).
  - *macOS*: `brew install juliaup` (or `curl -fsSL https://install.julialang.org | sh`).
  - *Ubuntu*: `curl -fsSL https://install.julialang.org | sh`.
  - Then `set_julia_path("~/.juliaup/bin/")` if Julia is not already on the `PATH`.
- **Python** — install [Miniforge](https://github.com/conda-forge/miniforge)
  (recommended; gives you `conda`).
  - *Windows*: run the `Miniforge3` installer; avoid the Microsoft Store "App
    execution alias" python — `en_check_python()` flags it as non-functional.
  - *macOS*: `brew install miniforge` (or the `arm64`/`x86_64` installer).
  - *Ubuntu*: `bash Miniforge3-Linux-x86_64.sh` from the
    [releases](https://github.com/conda-forge/miniforge/releases/latest).
- **GAMS** — install from <https://www.gams.com/download/> (license required),
  then point energyRt at the GAMS system directory:
  - *Windows*: `set_gams_path("C:/GAMS/<version>/")`.
  - *macOS*: `set_gams_path("/Library/Frameworks/GAMS.framework/Versions/Current/Resources")`.
  - *Ubuntu*: `set_gams_path("/opt/gams/gams<version>_linux_x64_64_sfx")`.

## 4. Auto-install the library layer

Once the runtimes are present, install the packages each backend needs in one
call:

```{r}
en_install_deps()   # detects runtimes, then installs Julia + Python + R libs
```

`en_install_deps()` runs a dependency check, installs only the safe library
layer for the runtimes it finds (skipping — with a warning — any runtime that is
missing), and re-checks at the end. You can also install per backend:

```{r}
en_install_julia_pkgs()      # JuMP, HiGHS, Cbc, Clp, ... into the Julia env
en_install_python_deps()     # conda env "energyRt" with pyomo + cbc (or pip)
en_install_python_deps(use_conda = FALSE)  # pip into the configured python
```

### Manual alternative

If a helper fails, run the same steps by hand.

R layer — the GAMS GDX bridge (not on CRAN) plus optional I/O helpers:

```r
pak::pkg_install("lolow/gdxtools")   # or remotes::install_github("lolow/gdxtools")
install.packages(c("jsonlite", "readxl", "openxlsx"))
```

Julia layer — from the Julia REPL (or `julia -e "..."`):

```julia
using Pkg
Pkg.add(["JuMP", "HiGHS", "Cbc", "Clp", "RData", "RCall",
         "CodecBzip2", "Gadfly", "DataFrames", "CSV", "SQLite", "Dates"])
```

Python/Pyomo layer — conda (recommended) or pip:

```bash
conda create -y -n energyRt python
conda install -y -n energyRt -c conda-forge pyomo pandas pyarrow coincbc
# pip fallback (no CBC binary): python -m pip install --upgrade pyomo pandas pyarrow
```

## 5. Point energyRt at the installations

If a runtime is not on your `PATH`, set its path (persisted in energyRt
options). For Python/Pyomo, point at the conda environment created above:

```{r}
set_julia_path("~/.juliaup/bin/")             # same on all platforms
set_python_path("~/.conda/envs/energyRt")     # or the env you created
set_glpk_path("C:/rtools45/x86_64-w64-mingw32.static.posix/bin")  # Windows (Rtools 4.5); unneeded on macOS/Ubuntu
set_gams_path("C:/GAMS/<version>/")           # see step 3 for macOS/Ubuntu paths
```

## 6. Verify end-to-end

Re-check (everything you installed should now be green), choose a default
solver, and solve a tiny model:

```{r}
en_check_dependencies(solver_pkgs = TRUE)   # also verifies JuMP/HiGHS in Julia
set_default_solver(solver_options$julia_highs_barrier)  # or solver_options$glpk

# a minimal model solve confirms the full toolchain works
# (see the "Model bricks" article for assembling a model)
```

If two or more backends are installed, solving the same scenario with each and
comparing objectives is the strongest check that your setup is correct.

## 7. Optional extras

Everything below is optional — energyRt solves models without any of it.

```{r}
# plotting (all autoplot()/draw() charts) + levcost side-by-side panels
install.packages(c("ggplot2", "patchwork"))

# report(): HTML datasheets & model/scenario reports need pandoc.
# RStudio ships pandoc; outside RStudio install it from pandoc.org
# (or reuse Quarto's: Sys.setenv(RSTUDIO_PANDOC = "<quarto>/bin/tools")).
install.packages("rmarkdown")

# report(format = "pdf" / "tex") additionally needs a LaTeX toolchain:
install.packages("tinytex")
tinytex::install_tinytex()

# maps for plot_trade_map() / the utopia$map layouts
install.packages("sf")

# GAMS gdx exchange
# install.packages("gdxtools", repos = "...")   # see the gdxtools README
```
