---
title: "Aggregating regions without averaging them away"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
# eval = FALSE: this article builds and solves a dozen models. The code is run
# and every number below checked out-of-band by
# dev/verify-region-aggregation.R; the outputs shown are those results.
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4
)
```

```{r setup}
library(energyRt)
library(geoscales)
library(clusterscales)
library(dplyr)
library(ggplot2)
```

# The problem with a weighted mean

A multi-region model gets expensive in the regions. The usual answer is to
aggregate: pick a coarser geoframe, roll the fine regions up into it, solve the
smaller problem. `aggregate_model_regions()` does exactly that -- extensive
quantities are summed, intensive ones take a capacity-weighted mean.

That is the right move when the regions are alike, and the wrong one when they
are not. A coal fleet built over forty years does not have *one* efficiency. Its
regional spread is the reason the model dispatches in a merit order at all, and
a mean erases it.

The spread does not have to be lost. A technology can carry **clusters** --
`technology@cluster` describes "a parallel sub-process of the same technology
... with its own capacity, availability and costs" -- so a zone-level technology
can keep several regional variants side by side. Aggregating then has a dial
rather than a switch:

| clusters per zone | what you get |
|---|---|
| one | `aggregate_model_regions()`: a single averaged technology |
| several | fewer variants than regions, but the spread that matters |
| one per region | nothing averaged at all |

This article builds one model, turns one thing at a time, and measures when the
dial matters. Two settings leave the answer untouched; three change it, and in
one of them the aggregated model comes out *cheaper* than the truth.

# The regions

TOPIA's eleven regions nest into three zones.

```{r regions}
gs <- topia$geoscales$honeycomb
regions <- geoscale_regions(gs, geoframe = "region")
geo <- as.data.frame(geoscale_leaftable(gs))

plot_geoscale(gs, type = "map")
```

# The plants

Seven archetypes of coal plant, each a contiguous block inside one zone.
Without real structure there is nothing for a clustering to find, and every
`k` looks equally bad.

```{r plants}
plants <- tibble::tribble(
  ~region, ~eff, ~invcost,     # efficiency, MEUR/GW
  "W1",    0.42,     1000,
  "W2",    0.33,     1500,
  "C1",    0.35,     1100,
  "C2",    0.35,     1100,
  "C3",    0.42,     1050,
  "C4",    0.42,     1050,
  "C5",    0.34,     1600,
  "C6",    0.34,     1600,
  "E1",    0.43,      980,
  "E2",    0.43,      980,
  "E3",    0.32,     1450
)
```

Real fleets are not that tidy, so the working model puts a ~1% jitter on both
columns. `E1a` below turns it off to isolate the exact case.

```{r jitter}
set.seed(42)
plants <- plants |>
  mutate(
    eff = round(eff * (1 + rnorm(n(), sd = 0.006)), 4),
    invcost = round(invcost * (1 + rnorm(n(), sd = 0.01)))
  )
```

# Load, and the siting limit

One load shape for every region (PJ per timeslice), scaled to a level that
rises west to east.

```{r demand}
calendar <- topia$modules$calendars$topia_seasons

shape <- topia$demand |>
  filter(calendar == "topia_seasons") |>
  left_join(
    as.data.frame(calendar@timetable) |> select(timeslice, share),
    by = "timeslice"
  ) |>
  mutate(w = load * share / sum(load * share))

demand <- plants |>
  mutate(annual = seq(80, 120, length.out = n())) |>     # PJ/year
  select(region, annual) |>
  tidyr::crossing(shape |> select(timeslice, w)) |>
  mutate(demand = annual * w) |>
  select(region, timeslice, demand)
```

Each plant is capped near its own region's peak. Without that cap the model
builds **three plants out of twenty-two**: once regions share a copperplate
their plants compete for one demand and the merit order keeps only the cheapest,
so competition erases the heterogeneity before any aggregation happens. The
honest reason for a cap is that you cannot put a zone's entire fleet on one
site.

```{r siting}
siting <- demand |>
  left_join(shape |> select(timeslice, share), by = "timeslice") |>
  group_by(region) |>
  summarise(cap.up = max(demand / share), .groups = "drop")   # GW
```

# Two technologies, eleven regions each

The region-varying parameters live in the slots, so this is **one** object per
fuel rather than one per region -- and `aggregate_model_regions()` clusters
either shape.

```{r techs}
ECOA <- newTechnology(
  name = "ECOA",
  desc = "Coal plant, one variant per region",
  region = plants$region,
  input = list(comm = "COA", unit = "PJ"),
  output = list(comm = "ELC", unit = "PJ"),
  ceff = plants |> transmute(region, comm = "COA", cinp2use = eff),
  invcost = plants |> select(region, invcost),
  fixom = plants |> transmute(region, fixom = 0.03 * invcost),
  capacity = siting,
  af = data.frame(af.up = 0.85),
  olife = data.frame(olife = 30)
)

EGAS <- newTechnology(
  name = "EGAS",
  desc = "Gas plant, one variant per region",
  region = plants$region,
  input = list(comm = "GAS", unit = "PJ"),
  output = list(comm = "ELC", unit = "PJ"),
  ceff = plants |> transmute(region, comm = "GAS", cinp2use = round(eff * 1.25, 4)),
  invcost = plants |> transmute(region, invcost = round(invcost * 0.6)),
  fixom = plants |> transmute(region, fixom = 0.03 * round(invcost * 0.6)),
  capacity = siting,
  af = data.frame(af.up = 0.85),
  olife = data.frame(olife = 30)
)

draw(ECOA)
```

# Transmission inside a zone

Adjacent regions in the same zone are linked, amply and losslessly. That makes
each zone a copperplate already -- which is the assumption a cluster
aggregation makes anyway, so here it is made explicitly rather than smuggled in
later. `E2a` removes it and measures what it was worth.

```{r links}
links <- gs |>
  geoscale_geometry(geoframe = "region") |>
  sf::st_touches() |>
  as.data.frame() |>
  transmute(src = regions[row.id], dst = regions[col.id]) |>
  filter(src < dst) |>
  left_join(geo |> select(src = region, src_zone = zone), by = "src") |>
  left_join(geo |> select(dst = region, dst_zone = zone), by = "dst") |>
  filter(src_zone == dst_zone) |>
  select(src, dst)

routes <- bind_rows(links, links |> rename(src = dst, dst = src))

TBD_ELC <- newTrade(
  name = "TBD_ELC",
  commodity = "ELC",
  routes = routes,
  trade = routes |> mutate(teff = 1.0),
  capacity = data.frame(stock = 1e4)             # GW: ample on purpose
)
```

# The model

```{r model}
COA <- newCommodity(name = "COA", unit = "PJ", timeframe = "ANNUAL")
GAS <- newCommodity(name = "GAS", unit = "PJ", timeframe = "ANNUAL")
ELC <- newCommodity(name = "ELC", unit = "PJ", timeframe = "HOUR")

SUP_COA <- newSupply(
  name = "SUP_COA", commodity = "COA", unit = "PJ",
  supply = data.frame(region = regions, cost = 2.0)      # MEUR/PJ
)
SUP_GAS <- newSupply(
  name = "SUP_GAS", commodity = "GAS", unit = "PJ",
  supply = data.frame(region = regions, cost = 6.0)
)
DEM_ELC <- newDemand(name = "DEM_ELC", commodity = "ELC", demand = demand)

repo <- newRepository(
  name = "agg",
  COA, GAS, ELC, SUP_COA, SUP_GAS, DEM_ELC, ECOA, EGAS, TBD_ELC
)

mod <- newModel(
  name = "AGG",
  region = regions,
  calendar = calendar,
  horizon = newHorizon(2025),
  discount = 0.05,
  data = repo
) |>
  setGeoscale(gs)

autoplot(mod)
```

# Solving it

```{r solve}
scen_fine <- mod |>
  interpolate_model(name = "FINE") |>
  solve_scenario(solver = solver_options$glpk)

getData(scen_fine, name = "vObjective", merge = TRUE)$value
#> [1] 185892.859193
```

GLPK is used throughout because every model here is small -- eleven regions and
four timeslices -- and at that size it is the quickest way to an answer. Do not
carry it to a real one. GLPK stops being usable well before the sizes that make
region clustering worth doing in the first place, so at any realistic
resolution solve with HiGHS through `julia` or `pyomo`:

```{r solver-note, eval = FALSE}
solve_scenario(scen, solver = solver_options$julia_highs)   # or $pyomo_highs
```

The clustering itself never solves anything -- `process_cluster_sweep()` and
`levcost()` are analytic -- so the solver only matters for the runs being
compared.

# What the plants cost, before any clustering

`levcost()` prices every process the model holds, analytically -- no solver.
Sorted, the eleven coal variants fall into the seven archetypes they were built
from, which is exactly the structure a clustering has to find.

```{r levcost}
levcost(mod, comm = "ELC", name = "ECOA", by_region = TRUE)
#> levcost_list (11)
#>   W1: 119.1202
#>   W2: 171.4738
#>   C1: 128.3743
#>   C2: 128.5894
#>   C3: 122.9501
#>   C4: 121.8456
#>   C5: 179.9411
#>   C6: 180.4450
#>   E1: 115.6380
#>   E2: 113.9066
#>   E3: 165.4407
```

`E1` is the cheapest archetype and `C6` the dearest -- 116 against 180
MEUR/PJ, a 56% spread across plants the model would otherwise average into
one. Priced region by region the eleven variants fall into seven steps:

| region | LCOE | | region | LCOE |
|---|---|---|---|---|
| `E2` | 113.9 | | `C2` | 128.6 |
| `E1` | 115.6 | | `E3` | 165.4 |
| `W1` | 119.1 | | `W2` | 171.5 |
| `C4` | 121.8 | | `C5` | 179.9 |
| `C3` | 123.0 | | `C6` | 180.4 |
| `C1` | 128.4 | | | |

Those seven steps are exactly what a clustering has to find, and `levcost()`
finds them analytically -- no solver, no scenario.

Note `by_region = TRUE`. levcost() builds a single-region mini-model, so
without it a technology spanning several regions is priced at just one of them
and a bare `levcost(mod, comm = "ELC")` reports one number for `ECOA` rather
than a ladder. `by_variant = TRUE` does the same along the other two axes,
vintage and cluster -- so a fleet with build years prices per vintage.

# Which processes may be merged

Only structurally identical ones. A cluster is a variant of the *same* process,
so a coal plant and a gas plant stay separate however alike their costs look.
The class leads the signature, so a supply never joins a technology that
happens to touch the same commodity:

```{r groups}
get_process_groups(mod)[, c("group", "class", "signature", "n")]
#>     group      class                signature  n
#> 1    ECOA technology in:COA|out:ELC|aux:|grp: 11
#> 2    EGAS technology in:GAS|out:ELC|aux:|grp: 11
#> 3 SUP_COA    supply         comm:COA|unit:PJ 11
#> 4 SUP_GAS    supply         comm:GAS|unit:PJ 11
#> 5     TRD      trade               TRD::ELC::  1
```

Five groups, not two. `supply`, `storage`, `import` and `export` carry a
`@cluster` slot as well, so a region grouping becomes sub-processes on them
too, and `n` counts the regions they span. `demand` is absent by design: it has
no `@cluster` slot, because a demand is a requirement rather than a process
with archetypes, and summing it over a zone is the right answer.

`TRD` is the odd one. A corridor has no `@region` slot -- its regions are
`src`/`dst` pairs -- so grouping it means partitioning *corridors* between a
pair of coarse regions, not clustering regions, and `n` counts corridors
rather than regions. Here there is one, so there is nothing to partition; the
[corridor section](#grouping-corridors) uses a model where there is.

The supplies here are flat -- every region pays 2.0 for coal -- so there is
nothing for a clustering to find, and asking for one says so rather than
inventing structure:

```{r groups-flat}
aggregate_model_regions(mod, geoscale = gs, level = "zone",
                        clusters = list(SUP_COA = 7))
#> Error: no usable clustering features: every candidate parameter is constant
#> or absent across the group's regions. Pass `features=`.
```

A supply whose cost really does vary by region is the case worth clustering:
plain aggregation gives a zone one capacity-weighted average price, while
clusters keep its cheap tranches separate from its dear ones -- the same
argument as for the plants, one layer upstream.

# The geometry constrains the clustering

Two regions may be merged only if they are adjacent, and adjacency drops every
edge that crosses a zone boundary:

```{r adjacency}
touching <- gs |>
  geoscale_geometry(geoframe = "region") |>
  sf::st_touches(sparse = FALSE)

sum(touching) / 2                                   # pairs on the map
#> [1] 18

zone <- setNames(geo$zone, geo$region)
sum(touching & outer(zone[regions], zone[regions], "==")) / 2
#> [1] 12
```

A cluster has to be connected in that graph, so it **cannot straddle a zone**,
and the graph has one component per zone -- which puts a floor under `k`.

# Settings are per family, not per model

`clusters` is keyed by technology group, and a single number for the whole
model is refused when there is more than one:

```{r k-scalar}
aggregate_model_regions(mod, geoscale = gs, level = "zone", clusters = 7)
#> Error: `clusters = 7` sets one number for all 5 process groups of this
#> model (ECOA, EGAS, SUP_COA, SUP_GAS, TRD), which is rarely what you want:
#> each family clusters on different parameters and has its own structure, so
#> it has its own k. Name them:
#>     clusters = list(ECOA = 7, EGAS = ...)
#>   Use get_process_groups() to see what can be merged, and
#>   process_cluster_sweep() to choose each k.
```

That is deliberate. A coal fleet groups on cost and efficiency, a wind fleet on
resource quality and the *shape* of its availability, and the right number
differs with both the spread and what the family is for. The error makes you
look at `get_process_groups()` before deciding.

Named, the floor shows itself:

```{r k-floor}
aggregate_model_regions(mod, geoscale = gs, level = "zone",
                        clusters = list(ECOA = 2, EGAS = 2))
#> Error: k = 2 is out of range for group 'ECOA': clusters cannot straddle a
#> `zone` region, so k runs from 3 (one cluster per `zone` region -- what
#> aggregate_model_regions() produces) to 11 (one per region, nothing averaged).
```

# Choosing k by looking

`process_cluster_sweep()` clusters one family over the whole admissible range
and reports what each `k` buys, over the same features and the same contiguity
graph the aggregation would use. Nothing is changed -- the point is to see
where the structure is before naming a number:

```{r sweep}
process_cluster_sweep(mod, "ECOA", geoscale = gs, level = "zone")
#>    k      within silhouette smallest                 sizes
#> 1  3 15.57484465 -0.1915882        2                 6,3,2
#> 2  4 12.05414650 -0.2325999        1               6,2,2,1
#> 3  5  8.66725287 -0.2220126        1             6,2,1,1,1
#> 4  6  3.29096595  0.1677981        1           4,2,2,1,1,1
#> 5  7  0.30748448  0.5461881        1         2,2,2,2,1,1,1
#> 6  8  0.15692375  0.4535683        1       2,2,2,1,1,1,1,1
#> 7  9  0.08150451  0.2968379        1     2,2,1,1,1,1,1,1,1
#> 8 10  0.01604168  0.1798995        1   2,1,1,1,1,1,1,1,1,1
#> 9 11  0.00000000  0.0000000        1 1,1,1,1,1,1,1,1,1,1,1
```

Read it as a scree plot. Dispersion falls with `k` and always will, so take the
knee rather than the minimum: `within` drops by an order of magnitude between
`k = 6` and `k = 7` and then only crawls, because seven is how many distinct
archetypes the data has -- past that, `k` splits members that differ by the
jitter alone. The silhouette peaks in the same place, and `sizes` says the
grouping is balanced there rather than one big cluster beside a row of
singletons.

`ggplot2` makes the knee easier to see than the table does:

```{r sweep-plot, fig.height = 3}
process_cluster_sweep(mod, "ECOA", geoscale = gs, level = "zone") |>
  select(k, within, silhouette) |>
  pivot_longer(-k) |>
  ggplot() +
  aes(x = k, y = value) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 7, linetype = "dashed") +
  facet_wrap(~name, scales = "free_y") +
  scale_x_continuous(breaks = 3:11) +
  labs(x = "clusters", y = NULL,
       title = "ECOA: dispersion and silhouette by k") +
  theme_bw()
```

`k = "auto"` applies a rule -- the best average silhouette over that sweep --
and hands the sweep back so the judgement can be checked rather than trusted.
`clusterscales` ships no `best_k()` deliberately: "the numbers inform a
judgement that stays yours." Sweep and look; use `"auto"` to reconnoitre, not
to decide.

```{r clusters}
mod_7 <- aggregate_model_regions(
  mod,
  geoscale = gs,
  level = "zone",
  clusters = list(ECOA = 7, EGAS = 7)
)

getObject(mod_7, name = "ECOA", drop = TRUE)@cluster
#>   cluster  desc  region order
#> 1   C1_C2 C1+C2 CENTRAL     1
#> 2   C3_C4 C3+C4 CENTRAL     2
#> 3   C5_C6 C5+C6 CENTRAL     3
#> 4   E1_E2 E1+E2    EAST     4
#> 5      E3    E3    EAST     5
#> 6      W1    W1    WEST     6
#> 7      W2    W2    WEST     7
```

`desc` records which regions each cluster came from, and `region` is the coarse
region it lives in -- a cluster exists only where its members did. A cluster of
one region takes that region's own code, so the 1:1 end reads `W1`, `C3`,
rather than `c01`.

Negative silhouettes at low `k` are normal here -- that is the contiguity
constraint, not a failure. With unstructured data the rule degenerates: the
silhouette climbs monotonically to `k = n` and "best" becomes "do not aggregate
at all". A clean peak means the structure is real.

# Seeing the clustering

A grouping you cannot look at is a grouping you cannot check.
`model_clusters()` hands back the crosswalk, the sweep, and a **geoscale
carrying a `cluster` geoframe** -- so the views that already exist show the
method with no new plotting code:

```{r cluster-map, fig.height = 4}
cl <- model_clusters(mod_7)

plot_geoscale(cl$ECOA$geoscale, type = "map", geoframe = "cluster")
```

The atoms are coloured by the cluster they joined. Every cluster is one
connected blob and none crosses a zone boundary -- that is the contiguity
constraint, visible rather than asserted.

```{r cluster-icicle, fig.height = 3}
plot_geoscale(cl$ECOA$geoscale, type = "icicle")
```

The icicle reads `nation > zone > cluster > region`: the cluster layer is an
extra level *inserted* between the target and the atoms, which is exactly what
a clustered aggregation is.

```{r cluster-leaftable}
as.data.frame(geoscale_leaftable(cl$ECOA$geoscale))
#>    nation    zone cluster region
#> 1   TOPIA    WEST      W1     W1
#> 2   TOPIA    WEST      W2     W2
#> 3   TOPIA CENTRAL   C1_C2     C1
#> 4   TOPIA CENTRAL   C1_C2     C2
#> 5   TOPIA CENTRAL   C3_C4     C3
#> 6   TOPIA CENTRAL   C3_C4     C4
#> 7   TOPIA CENTRAL   C5_C6     C5
#> 8   TOPIA CENTRAL   C5_C6     C6
#> 9   TOPIA    EAST   E1_E2     E1
#> 10  TOPIA    EAST   E1_E2     E2
#> 11  TOPIA    EAST      E3     E3
```

This geoscale is a **view**, not the model's own. The aggregated model's
geoscale is pruned to `zone`, and its atoms have to keep matching its regions or
`interpolate_model()` misreads what resolution it is looking at.

# Two families, two settings

Both families have the same eleven regions and the same contiguity graph, so it
is tempting to give them one `k`. Their sweeps even agree -- here, because
`EGAS` was built as a scaled copy of `ECOA`, so its parameters spread the same
way:

```{r sweep-both}
process_cluster_sweep(mod, "EGAS", geoscale = gs, level = "zone") |>
  select(k, within, silhouette) |>
  left_join(
    process_cluster_sweep(mod, "ECOA", geoscale = gs, level = "zone") |>
      select(k, within_coa = within, sil_coa = silhouette),
    by = "k"
  )
#>    k      within silhouette  within_coa     sil_coa
#> 1  3 15.57976856 -0.1919701 15.57484465 -0.1915882
#> 2  4 12.06067920 -0.2329213 12.05414650 -0.2325999
#> 3  5  8.67779012 -0.2236994  8.66725287 -0.2220126
#> 4  6  3.29457713  0.1649621  3.29096595  0.1677981
#> 5  7  0.31202956  0.5430788  0.30748448  0.5461881
#> 6  8  0.15829511  0.4519464  0.15692375  0.4535683
#> 7  9  0.07961517  0.2960823  0.08150451  0.2968379
#> 8 10  0.01433427  0.1801032  0.01604168  0.1798995
#> 9 11  0.00000000  0.0000000  0.00000000  0.0000000
```

And yet the same `k` is the wrong choice for both, because what a clustering
costs is not how the parameters spread -- it is what the model *does* with the
difference. Four runs settle it:

```{r two-settings}
aggregate_model_regions(mod, geoscale = gs, level = "zone",
                        clusters = list(ECOA = 7, EGAS = 3)) |>
  interpolate_model(name = "K73") |>
  solve_scenario(solver = solver_options$glpk) |>
  getData(name = "vObjective", merge = TRUE) |>
  pull(value)
#> [1] 186705.498894
```

| `ECOA` | `EGAS` | objective | gap to fine |
|---|---|---|---|
| 11 | 11 | 185892.859193 | -- (the fine model) |
| 7 | 7 | 186063.417050 | +0.0918% |
| 7 | 3 | 186705.498894 | +0.4372% |
| 3 | 7 | 192627.056726 | +3.6226% |
| 3 | 3 | 193269.138570 | +3.9680% |

Read the last two rows against each other. Coarsening **gas** alone costs
0.35 points; coarsening **coal** alone costs 3.53 -- ten times as much, from the
same `k`, on the same regions, with the same spread in the data. Coal is on the
margin, where a fraction of a point of efficiency decides which plant runs; gas
sits at its siting ceiling in every zone, so averaging its cost across a zone
changes almost nothing. One `k` for the model either wastes variants on gas or
ruins coal.

Which family deserves the resolution is an empirical question about the model,
not a property of the data, and no rule reads it off the sweep. That is why
`clusters` is keyed by family and why a bare number is refused.

`features` is per family for the same reason. It takes the parameter names to
cluster on, or a **function** of the member technologies returning one row per
region -- which is how a *profile shape* enters, since scalars cannot see shape.
`E2c` below is that case: there the clustering has to choose one availability
profile per cluster, and it takes a real member's rather than a fabricated mean.

# Plain aggregation, and the same thing said with clusters

`clusters = NULL` is the plain aggregation: each zone collapses to one value per
parameter. On this model that is *exactly* `k = 3` -- one cluster per zone,
nothing left to distinguish:

```{r plain-vs-3}
aggregate_model_regions(mod, geoscale = gs, level = "zone") |>
  interpolate_model(name = "PLAIN") |>
  solve_scenario(solver = solver_options$glpk) |>
  getData(name = "vObjective", merge = TRUE) |>
  pull(value)
#> [1] 193269.138570
```

Identical to the `3/3` row above, to the last digit. The floor under `k` is the
plain aggregation, and the two are the same model written two ways -- so
`clusters` is an argument you turn up from there, not a different function.

# Comparing runs, not just cost

The runs are not keyed alike: the fine model reports eleven regions, a clustered
run three zones and one variant per cluster (`ECOA_CLC1C2` ...). No name surgery
is needed for either, because solved output carries the object a variant came
from in `base`, and its cluster in `cluster`. The leaftable maps region to zone.

```{r align}
getData(scen_fine, name = "vTechCap", merge = TRUE) |>
  mutate(zone = geo$zone[match(region, geo$region)]) |>   # W1 -> WEST
  summarise(capacity = sum(value), .by = c(zone, base)) |>
  arrange(zone, base)
#>      zone base capacity
#> 1 CENTRAL ECOA  191.123
#> 2 CENTRAL EGAS 1083.031
#> 3    EAST ECOA  113.114
#> 4    EAST EGAS  640.978
#> 5    WEST ECOA   53.306
#> 6    WEST EGAS  302.070
```

A clustered run reports zones already, so there `zone` is `region` and nothing
but the `.by` changes. For the general case `compare_scenarios()` does this
properly -- overview, per-variable difference counts and the largest absolute
deltas -- once the keys are aligned.

# Solving at the top level

There is no way to solve a fine model at a coarse level *without* aggregating.
`interpolate_model()` sorts a geoscale by how its atoms line up with the model's
regions, and a geoscale whose atoms are coarser than the model is refused with a
pointer back here. `prune_geoscale()` is what this function calls to set the
result's geoscale -- it is not an alternative route. The two real routes upward
are to aggregate a fine model, as below, or to declare the model coarse from the
start, which `vignette("space-resolution")` demonstrates.

TOPIA's top frame is `nation`, one region:

```{r nation}
mod_nation <- aggregate_model_regions(
  mod,
  geoscale = gs,
  level = "nation",
  clusters = list(ECOA = 11, EGAS = 11)
)

mod_nation@config@region
#> [1] "TOPIA"
```

| level | clusters | objective | gap to fine |
|---|---|---|---|
| region | -- | 185892.859193 | -- (the fine model) |
| zone | 7 / 7 | 186063.417050 | +0.0918% |
| nation | 11 / 11 | 184380.820211 | **-0.8134%** |
| nation | plain | 193259.899661 | +3.9631% |

The third row is the one to look at. Eleven clusters over one region keeps
**every** parameter -- nothing is averaged, the 1:1 case -- and still comes out
0.81% **cheaper** than the model it came from. Nothing about the plants changed;
the topology did. The fine model links regions only inside a zone, and
collapsing the three zones into one nation hands it three copperplates joined
into one, so `EAST`'s cheap coal can serve `WEST`'s load over a line that does
not exist. `E2a` below is the same mechanism measured on purpose.

That is the general rule for coarsening: a 1:1 clustering proves the
**parameters** survived, never that the aggregation was safe. Whatever the
regions were keeping apart is gone as soon as they share a balance.

# Sampling a sub-territory

Coarsening the whole model and taking part of it at full resolution are
different moves. `subset_model_regions()` keeps the model's own regions and
drops the rest, so it composes with everything above -- sample first, then
cluster what is left:

```{r subset}
mod_sub <- subset_model_regions(
  mod,
  region = geo$region[geo$zone %in% c("WEST", "CENTRAL")]
)
#> spatial sample: kept 8 region(s); dropped 0 object(s) and 6 trade route(s)
#>   outside the sample (verbose = TRUE lists them)

get_process_groups(mod_sub)[, c("group", "n")]
#>   group n
#> 1  ECOA 8
#> 2  EGAS 8
```

The families are the same families, eight regions deep instead of eleven, and
the admissible `k` moves with them -- two zones now, so the floor is 2 and the
ceiling 8:

```{r subset-sweep}
process_cluster_sweep(mod_sub, "ECOA", level = "zone")
#>   k      within silhouette smallest           sizes
#> 1 2 12.23946641 -0.1520565        2             6,2
#> 2 3  8.74669572 -0.4145140        1           6,1,1
#> 3 4  3.48880116  0.0952621        1         4,2,1,1
#> 4 5  0.16577247  0.6206045        1       2,2,2,1,1
#> 5 6  0.08257609  0.4081067        1     2,2,1,1,1,1
#> 6 7  0.01674316  0.2475210        1   2,1,1,1,1,1,1
#> 7 8  0.00000000  0.0000000        1 1,1,1,1,1,1,1,1
```

The knee is at `k = 5` here, not 7: `EAST` took two of the eleven archetypes
with it. The sub-territory's structure is its own, and `k` chosen on the full
model would be wrong for it.

```{r subset-solve}
aggregate_model_regions(mod_sub, level = "zone",
                        clusters = list(ECOA = 5, EGAS = 5)) |>
  interpolate_model(name = "SUB5") |>
  solve_scenario(solver = solver_options$glpk) |>
  getData(name = "vObjective", merge = TRUE) |>
  pull(value)
#> [1] 130655.145340
```

Against the sampled fine model's 130572.373965 that is +0.0634% -- the same
order as `7/7` on the whole model, for the same reason: five clusters is the
resolution these eight regions actually have.

## One thing sampling can change under you

A cost declared at a **coarse** cell is charged once, in full, at that cell --
it does not know how many of the cell's children are still in the model. Sample
away some of them and the sub-model carries the whole corridor's cost, so its
objective stops being comparable with the full model's. `subset_model_regions()`
says so:

```{r subset-warn}
TBD_PRICED <- newTrade(
  name = "TBD_ELC", commodity = "ELC", routes = routes,
  trade = routes |> mutate(teff = 1.0),
  capacity = data.frame(stock = 1e4),
  invcost = data.frame(region = "CENTRAL", invcost = 50)   # a ZONE, not a reg.
)

mod_priced <- mod |>
  add(TBD_PRICED, overwrite = TRUE) |>     # replaces the unpriced corridor
  subset_model_regions(region = c("C1", "C2", "C3"))
#> Warning: Sampling regions changes a trade cost declared at a coarse level:
#>   TBD_ELC@invcost at 'CENTRAL' (keeps 3 of 6 region(s): dropped C4, C5, C6)
#>   A coarse cost is charged ONCE at that cell, whatever is left under it, so
#>   this sub-model bears the WHOLE corridor cost rather than the share its
#>   remaining endpoints should -- its objective is not comparable with the full
#>   model's.
#>   Declare the cost on the endpoint regions instead, so it follows the ones
#>   that survive.
```

It is silent when the whole cell survives, and silent for endpoint-named and
unregioned costs, which follow the endpoints that are left. Only the coarse form
has the problem, and naming the endpoint regions is the fix.

# Does aggregation change the answer?

Five experiments, each solving the fine eleven-region model, aggregating it to
the three zones, and comparing.

## E1a -- members identical: exact

With no jitter every member of an archetype has the *same* efficiency and
invcost, so the capacity-weighted mean is that value and clustering at `k = 7`
is lossless by construction. Objective identical to the last digit, capacity and
generation identical.

## E1b -- a realistic spread: near-exact

With the jitter back, members differ by about 1%. Clustering what is genuinely
alike costs the spread you chose to ignore and nothing else.

## E2a -- pooling grants transmission that does not exist

Drop `TBD_ELC` and each region must serve its own load, so all plants build.
Aggregating to zones hands the model a copperplate it never had.

```{r e2a}
mod_isolated <- newModel(
  name = "AGG_ISOLATED",
  region = regions,
  calendar = calendar,
  horizon = newHorizon(2025),
  discount = 0.05,
  data = newRepository("agg_isolated",
                       COA, GAS, ELC, SUP_COA, SUP_GAS, DEM_ELC, ECOA, EGAS)
) |>
  setGeoscale(gs)
```

Every aggregation comes out **cheaper** than the model it came from, at every
`k` including 1:1. That is the lesson: 1:1 clusters preserve the parameters
perfectly and the topology not at all. Capacity and generation do not move --
both are pinned by the siting cap and the demand -- but fuel burn drops about
9%, and with it emissions. A modeller reporting capacity sees nothing; one
reporting emissions is wrong by a tenth.

## E2b -- merging across the real structure

The last two experiments need quantities chosen by economics rather than pinned
by a ceiling, which takes one change: gas fuel at 12 instead of 6, at which
point gas sits at 78% of its cap rather than 100%.

```{r e2b}
SUP_GAS_12 <- newSupply(
  name = "SUP_GAS", commodity = "GAS", unit = "PJ",
  supply = data.frame(region = regions, cost = 12.0)
)
mod_economic <- newModel(
  name = "AGG_ECONOMIC",
  region = regions,
  calendar = calendar,
  horizon = newHorizon(2025),
  discount = 0.05,
  data = newRepository("agg_economic",
                       COA, GAS, ELC, SUP_COA, SUP_GAS_12, DEM_ELC,
                       ECOA, EGAS, TBD_ELC)
) |>
  setGeoscale(gs)
```

At `k = 3` dissimilar archetypes collapse into one average plant, coal capacity
**more than halves** and gas takes over -- because the averaged coal plant is
worse than the good half of the fleet it replaced. `k = 7`, clustering *along*
the structure, stays near-exact. The error is not aggregation as such; it is
merging things that are not alike.

## E2c -- one profile per cluster

Give each archetype an availability profile peaking at a different slice of the
day, so members differ in *shape* and not only in scalars. A profile is a
**link** to another object rather than a number, so it cannot be averaged:
`aggregate_model_regions()` gives each cluster its **medoid** member's profile --
a real member's shape rather than a fabricated mean. That is the better of the
two available answers and still lossy, because the other members' shapes are
gone. Averaging them would be worse: the mean of two offset peaks is flatter
than either, and the model would believe in availability no site ever had.

## The five together

| experiment | changed | mechanism | objective | quantities |
|---|---|---|---|---|
| E1a | no jitter | members identical | **exact** | identical |
| E1b | -- | ~1% spread | +0.0918% | identical |
| E2a | no `TBD_ELC` | pooling adds transmission | **-3.89%** | fuel -9% |
| E2b | gas 12, `k = 3` | merged across structure | **+5.07%** | coal cap -55% |
| E2c | profiles, `k = 3` | one profile per cluster | **+4.16%** | coal gen -48% |

Read the last two columns rather than the middle one. E2a is the only case where
the objective moves *down*, and the only one no choice of `k` fixes -- it is a
topology error, not a clustering error. E2b and E2c are clustering errors, and
both vanish at `k = 7`, the resolution the data actually has.

# What the 1:1 case does and does not prove

At `k = 11` the aggregated model reproduces the fine objective to the last digit
in E1a, E1b, E2b and E2c. That is a real check -- every parameter survived the
merge, extensive quantities were summed and intensive ones were not, and no
variant was silently dropped.

E2a keeps it honest. There `k = 11` is 3.9% cheaper than the model it came from
with every parameter intact, because the transmission limits did not survive the
aggregation -- and the `nation` run above is the same thing at 0.81% on the base
model, where every parameter survives and three zones merge into one. Use the
1:1 invariant to test the *parameter* carry-over, never as evidence that the
aggregation as a whole is safe.

# Caveats

* **Clusters are not regions.** They share one balance, one demand and one set
  of trade routes. Everything that made the regions distinct *other than* the
  clustered parameters is gone -- which is what E2a measures.
* **The weights matter.** Intensive parameters are averaged with one size per
  region, taken from capacity. A group whose members declare no capacity gets
  equal weights, which is rarely what you want.
* **`k = "auto"` is a rule, not an answer.** Sweep with
  `process_cluster_sweep()` and read it. If the silhouette has no peak, the
  regions do not cluster.
* **One `k` is not a model-wide setting.** Which family deserves the
  resolution is an empirical question -- here coarsening coal costs ten times
  what coarsening gas costs -- so `clusters` is keyed by family and a bare
  number is refused when there is more than one.
* **Only structurally identical technologies merge.** If a family should be
  merged but is not, compare the signatures -- an aux commodity in one member is
  enough to split it.
* **Check the quantities, then check why.** If they do not move, find out
  whether the aggregation is good or a constraint is doing the choosing. In the
  base model gas builds to exactly 100% of its siting cap in all three zones,
  which pins the build whatever the parameters say.
