---
title: "Reports and levelized costs"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  warning = FALSE,
  eval = requireNamespace("rmarkdown", quietly = TRUE) &&
    rmarkdown::pandoc_available("2.0")
)
```

energyRt renders **reports** — self-contained HTML/PDF/Word datasheets — for
technologies and storages, whole models and repositories, solved scenarios,
and scenario comparisons. Reports and the **levelized costs** (`levcost()`)
that feed them are *derived artifacts*: both are cached on disk and
re-computed only when their inputs change. This article tours the layer —
what renders where, how the caching works, and how to compare scenarios.

```{r setup, message = FALSE}
library(energyRt)
```

Everything lands under a throwaway project here; `open_project()` anchors
the registry, the object stores, and the two derived-artifact folders
(`reports/`, `levcosts/`) in one call:

```{r project}
proj <- file.path(tempdir(), "rep-demo")
dir.create(proj, showWarnings = FALSE)
open_project(proj)
```

A technology to price and report on:

```{r tech}
PP <- newTechnology(
  "PP", desc = "Gas power plant",
  input  = list(comm = "GAS", unit = "GWh"),
  output = list(comm = "ELC", unit = "GWh"),
  ceff   = data.frame(comm = "GAS", cinp2use = 0.55),
  invcost = data.frame(invcost = 800),
  fixom = data.frame(fixom = 20), varom = data.frame(varom = 2),
  af = data.frame(af.up = 0.85),
  vintage = data.frame(olife = 25L), cap2act = 8.76)
```

# Levelized costs, cached

`levcost()` prices a technology (LCOE), storage (LCOS), or trade route
(LCOT). `method = "auto"` uses a closed-form engine when the process
qualifies and falls back to a small solver run otherwise:

```{r levcost}
lc <- levcost(PP, fuel_costs = c(GAS = 25), discount = 0.06,
              base_year = 2025, verbose = FALSE)
round(as.numeric(lc$levcost_npv), 2)
```

The result was also written to the on-disk cache, keyed by the
technology's *content hash* plus the assumptions. Re-pricing the identical
question returns instantly from disk:

```{r levcost-cache}
lc2 <- levcost(PP, fuel_costs = c(GAS = 25), discount = 0.06,
               base_year = 2025)
list.files(get_levcost_cache_path())
```

Anything that changes the answer — the technology's data, `discount`, fuel
prices, the method — is part of the cache key, so a stale hit is
impossible; `force = TRUE` recomputes regardless, and `cache = FALSE`
skips the cache entirely. For objects *saved* in a store the cache moves
into the owning folder (`<scenario>/levcost/`, a model store entry's
`levcost/`); the project-level `levcosts/` above is the tier for in-memory
objects.

# Rendering a datasheet

`report()` renders a technology datasheet — parameters, the `draw()`
diagram, and (with `levcost =`) the levelized-cost section. The same rule
places the output: saved objects render inside their own folder's
`reports/`, in-memory objects into the project-level `reports/` tier:

```{r report-tech}
f <- report(PP, levcost = lc, format = "html", open = FALSE)
basename(f)
```

A report is **skip-if-current**: the rendered file carries a sidecar
(`*.report.yml`) keying the object content, the template content, the
arguments, and the levcost identity. Repeating the call re-renders
nothing:

```{r report-uptodate}
f <- report(PP, levcost = lc, format = "html", open = FALSE)
```

Editing the technology, the template, or any argument invalidates the key;
`force = TRUE` re-renders unconditionally. Formats: `format =` takes any of
`"html"`, `"pdf"`, `"tex"`, `"docx"` (PDF needs a LaTeX toolchain).

# Model and scenario reports

Container reports split by role: `report(model)` documents the
**assumptions** (configuration, inventories, per-process sheets), while
`report(scenario)` documents the **results** of a solved scenario — status,
generation/capacity charts, costs, emissions; `run =` renders any recorded
run, and `template = "full"` switches to the comprehensive page-broken
layout with maps and branding. Reporting a single process from a container
injects its container-priced levelized cost:

```{r report-scen, eval = FALSE}
report(scen)                       # results overview -> <scenario>/reports/
report(scen, run = "julia_highs")  # another recorded run
report(scen, template = "full")    # comprehensive, page-broken layout
report(scen, name = "PP")          # one process, priced in this scenario
```

The shipped templates are listed by `report_templates()`; custom `.Rmd`
templates build on the exported `report_*` helper family and are passed via
`template = "<path>"`.

# Element datasheets

Every model component has a `report()` method — commodities, supply,
demand, trade, import/export, weather, taxes and subsidies, user
constraints, calendars, horizons, configurations, and user costs each
render a one-object datasheet from their own shipped template:

```{r report-elements, eval = FALSE}
report(COA)                          # a commodity datasheet
report(SUP_COA)                      # a supply datasheet
report(mod, name = "SUP_COA")        # the same, found inside a model
report(mod, name = "X", class = "demand")   # narrow a name lookup
```

An object can carry its own default template in `misc$report` (also the
techspec `report:` key) — a template name or a file path; an explicit
`template = ` argument always wins:

```{r report-misc, eval = FALSE}
SUP_COA@misc$report <- "my_supply_sheet.Rmd"
report(SUP_COA)                      # renders my_supply_sheet.Rmd
```

# Comparing scenarios

`compare_scenarios()` takes a named list of solved scenarios (or one
scenario plus `runs =` to compare its recorded runs) and returns a
classed comparison — overview, per-variable deltas against the base, and
the largest movers; `autoplot()` charts it and `report()` renders the
comparative report:

```{r compare, eval = FALSE}
cmp <- compare_scenarios(list(BASE = base, CO2CAP = capped))
cmp                                # printed overview + top deltas
autoplot(cmp)                      # faceted comparison charts
report(cmp, open = FALSE)          # the comparative report (report_scenarios)
```

`compare_models()` diffs two models' declarations (inventory by content
hash, config, then per-slot drill-down), and `compare_inputs()` diffs two
interpolated scenarios' parameter stores — the tool for "same model,
different numbers?" questions. For quick ad-hoc comparisons, plain
`getData()` across scenarios remains the one-liner (see the *Workflow*
article):

```{r getdata-compare, eval = FALSE}
getData(c("BASE", "CO2CAP"), name = "vEmsFuelTot", comm = "CO2",
        merge = TRUE)
```

# Where everything lands

```
<project>/
├── reports/              # reports of in-memory objects (temporary tier)
├── levcosts/             # levcost cache of in-memory objects
├── scenarios/<scen>/
│   ├── reports/          # this scenario's reports
│   └── levcost/          # its levcost cache
└── models/<name>/
    ├── reports/          # a stored model's reports
    └── levcost/          # its levcost cache
```

Both tiers are *caches of derived artifacts*: safe to delete at any time,
rebuilt on demand, and never consulted for anything but skipping work.
