---
title: "Scenario management: how models, scenarios, and solutions are organized"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE
)
```

An energy-modeling project quickly accumulates artifacts: models built from
shared repositories, scenarios interpolated from them at different scopes,
and solutions produced by different solvers. energyRt organizes all of it
in one project folder with three principles:

1. **One folder per thing, one manifest per folder.** Every saved model,
   repository, scenario, and solver run is a directory with a small YAML
   manifest describing it. The manifests — not any central database — are
   the source of truth; the project registry is a rebuildable index over
   them.
2. **Share by content, not by copy.** Models, repositories and datasets
   are stored once, under their **name** (`models/SM/`); a content hash in
   the manifest detects unchanged saves and *verifies* references (other
   objects reference by name + hash). Embedding a private copy remains
   available where portability matters — "interlinking to avoid
   duplication is an option, not a rule".
3. **Derive paths, don't trust stored ones.** Where a run's files live is
   always computed from the scenario's location and the run's identity.
   Loading a scenario re-bases every stored path onto the folder it is
   actually loaded from, so moving or renaming folders — or opening a
   project from a different working directory — just works.

This article walks the whole layout bottom-up. The *Workflow* article shows
the same machinery in day-to-day use; this one explains how the pieces fit.

```{r setup}
library(energyRt)
```

Everything lands under a throwaway project folder here; in a real project
these are relative paths at the project root (the defaults are
`scenarios/`, `models/`, `repositories/`, and `energyRt_registry.csv`).

```{r paths}
proj <- file.path(tempdir(), "sm-demo")
set_scenarios_path(file.path(proj, "scenarios"))
set_models_path(file.path(proj, "models"))
set_repositories_path(file.path(proj, "repositories"))
set_registry_file(file.path(proj, "energyRt_registry.csv"))
```

A small model to work with — one region, one technology, three years:

```{r model}
cal <- newCalendar(
  timetable = make_timetable(
    struct = list(ANNUAL = "ANNUAL", HOUR = sprintf("t%02d", 1:4))),
  name = "smc")

make_mod <- function(stock) {
  TECH <- newTechnology(
    "TECH", output = data.frame(comm = "ELC", unit = "GWh"), cap2act = 1,
    capacity = data.frame(region = "R1", year = 2020:2022, stock = stock),
    fixom = data.frame(fixom = 1),
    vintage = data.frame(olife = 50L), optimizeRetirement = FALSE)
  newModel(
    name = "SM", region = "R1", discount = 0, calendar = cal,
    horizon = newHorizon(2020:2022, intervals = c(1, 1, 1)),
    repo = newRepository(
      "smr", newCommodity("ELC", timeframe = "HOUR"), TECH,
      newSupply("BACK", commodity = "ELC", supply = data.frame(cost = 1000)),
      newDemand("DEM", commodity = "ELC", demand = data.frame(demand = 1))))
}
mod <- make_mod(stock = c(100, 100, 100))
```

# The scenario folder

A **scenario** is one interpolated instance of a model: the model compiled
onto a concrete calendar, horizon, and region scope. Solving it — possibly
several times, with several solvers — produces **runs**. On disk:

```
scenarios/<name-model-calendar-horizon>/
├── scenario.yml        # manifest: model reference, calendar, horizon,
│                       #   `default:` — the default run (edit it freely)
├── scen.RData          # the thin S4 shell (all big data lives beside it)
├── class · format · layout · logfile.csv
├── modInp/parameters/<param>/           # the (base) interpolated problem
├── model/…             # the model's data — only when embedded (see below)
└── runs/               # one folder per solver run
    ├── glpk/                            # a base-problem run
    │   ├── run.yml     #   provenance: solver, cmdline, started/finished,
    │   │               #   duration, status, objective, host, user
    │   ├── solver/     #   the solver working dir; the solver writes its
    │   │               #   own solver/output/ — never restructured
    │   └── modOut/variables/<var>/      #   the parsed solution store
    └── julia_highs/…                    # another backend, side by side
```

Three ideas to notice:

* **A run owns its artifacts.** Solving the same scenario with a second
  backend does not overwrite anything: each run keeps its inputs, raw
  outputs, provenance, and parsed solution. The in-memory scenario object
  holds ONE active solution at a time; switching is a read, not a re-solve.
* **The run's location is derived.** Nothing stores "where run X lives" —
  it is always `<scenario>/runs/<run-id>/`, computed on demand. That is why
  a moved scenario folder cannot strand its runs.
* **`default:` in scenario.yml is yours.** It names the run a bare
  `read_solution(scen)` resolves to after loading; a save sets it to the
  active run, and you may edit the file to point elsewhere.

Let's produce that tree. `run = "<label>"` names a run explicitly; without
it, the label defaults to the solver directory name (e.g. `"glpk"`):

```{r solve}
scen <- interpolate_model(mod, name = "BASE") |>
  solve_scenario(solver = solver_options$glpk, echo = FALSE)
scen <- solve_scenario(scen, solver = solver_options$glpk, echo = FALSE,
                   force = TRUE, run = "glpk-alt")
scenario_runs(scen)[, c("run", "status", "solver_name", "objective", "active")]
```

`scenario_run_info(scen, "glpk")` returns one run's full `run.yml`;
`read_solution(scen, run = "glpk")` switches the active solution;
`drop_scenario_run(scen, "glpk-alt")` deletes a run (the active one only
with `force = TRUE`). An *external* working directory — outside the
scenario, no run record — is available via
`solve_scenario(solver.dir = "...")`, and `transient = TRUE` solves in a
throwaway folder.

```{r save}
scen <- save_scenario(scen, verbose = FALSE)
sort(list.files(scen@path))
```

The folder *names* — the `<name-model-calendar-horizon>` slug, store entry
basenames, the default solve label — are derived by built-in functions you
can override with `set_path_builder()` (see the toolbox below). Custom
names are safe because nothing is ever parsed back out of a folder name:
scenarios and store entries are found by the name recorded in their
manifests and the registry. The run *tree* (`runs/<variant>/<solve>/`) is
structural and not overridable.

# Models and repositories: stored once, verified by hash

The model is part of the scenario, and by default a saved scenario embeds
its own copy — self-contained, but wasteful when fifty scenarios share one
model. The **model store** removes the duplication:

```{r model-store}
mod <- save_model(mod, verbose = FALSE)      # models/SM/
scen <- save_scenario(scen, verbose = FALSE) # now saved with a REFERENCE
yaml::read_yaml(file.path(scen@path, "scenario.yml"))$model
```

The store folder is simply the model's **name** — `models/SM/` — and it
updates *in place*: edit a parameter, re-save, same folder. The content
hash (`model_hash()`, computed with all volatile bookkeeping stripped)
lives in the manifest and does the two jobs a hash is for: saving
*identical* content twice is a no-op, and references *verify* — a scenario
records the `{name, hash}` it was built against, and if the stored model
has changed since, loading warns loudly that results may not reproduce
(sealing, below, prevents exactly that; and
`save_model(rehash = FALSE)` keeps the recorded hash through a change you
declare insignificant — a version tag rather than a content proof, marked
`hash_kept:` in the manifest).
`load_scenario()` resolves the reference back through the registry and
errors — naming the fix — if the store is gone.
`save_scenario(embed_model = TRUE)` keeps a folder self-contained instead.

The hash logic in five lines:

```{r hash-logic, eval = FALSE}
model_hash(mod)                    # content identity (misc/paths stripped);
                                   #   repository_hash(), dataset_hash(),
                                   #   object_hash() for anything else
save_model(mod)                    # same hash as stored -> silent no-op
save_model(mod2)                   # changed -> SAME folder, new hash
save_model(mod2, rehash = FALSE)   # changed, but KEEP the old hash
load_model("UTOPIA_R11@a1b2c3d4")  # pin a version: verified against the
                                   #   manifest, errors if superseded
```

A reference whose recorded hash no longer matches the store loads the
*current* version with a loud warning ("results may not reproduce");
sealing (below) makes that warning impossible for finished work, because
sealed content cannot drift.

One level down, the same pattern applies to **repositories** — the shared
building-block collections that many models have in common:

```{r repo-store}
repo <- mod@data[[1]]
save_repository(repo, verbose = FALSE)        # repositories/smr/
mod2 <- save_model(make_mod(c(100, 100, 100)), overwrite = TRUE,
                   verbose = FALSE)           # repo now saved as a reference
```

`save_model(embed_repos = )` mirrors `embed_model`: with the repository in
its store, the model's entry records a `{name, hash}` reference instead of
a data copy, and `load_model()` re-assembles the chain — a scenario
referencing a model whose repositories are themselves references loads
back fully resolved.

One level further down sits the **dataset store** (`datasets/`), for the
data that dominates size and rarely changes: a weather or demand time
series, a geoscale map. `save_dataset(x, "name")` stores it once,
content-addressed like models and repositories; `save_repository()`,
`save_model()` and `save_scenario()` then reference it (`embed_datasets =`,
same `NULL`/`TRUE`/`FALSE` contract as `embed_repos`) instead of re-saving
the table with every model or scenario version. A dataset can also record a
*generating call* (`save_dataset(name = , fun = "pkg::fun", args = )`) with
a materialized snapshot, so "the data" and "the code that made it" are the
same store entry. Loading any owner resolves its dataset references back
automatically — see `?save_dataset`.

# The registry

Every save also records its object in the project **registry** — one CSV at
the project root indexing models, repositories, scenarios, and runs:

```{r registry}
reg <- load_registry()
reg[, c("type", "name", "parent")]
```

The registry is an *index, never an authority*: `refresh_registry()`
rebuilds it at any time by rescanning the stores and reading the manifests.
`find_in_registry(reg, type = "model", hash = "a1b2")` looks entries up by
type, name, (short) hash — or, for run rows, by `variant =`: a run's `name`
holds the plain solve label, its problem sits in the `variant` column, and
`parent` names the scenario (the worked example below shows a full table).

The registry is also what makes NAME-based access work. `getScenario("BASE")`
resolves the name, loads the thin on-disk shell, and caches it in the
session bank `.scen` — re-saved scenarios reload automatically, repeat
access is free. `getData()` accepts names directly, so comparing registered
scenarios needs no explicit loading at all:

```{r getdata-by-name}
getData("BASE", name = "vObjective", merge = TRUE)[, c("scenario", "value")]
```

`getModel()`, `getRepository()` and `getDataset()` are the typed
companions; `load_scenarios()` returns many scenarios as a `getData()`-ready
named list; and `open_project("path/")` points a session at another
project's registry and stores in one call.

# Variants: several problems in one scenario

A scenario sometimes needs *the same model asked a different question* —
a sampled calendar, a restricted horizon, a myopic step with carried-over
capacity. These are **own-problem variants**: named folders under `runs/`
that carry their *own* interpolated problem, with their solves inside.

```{r variant}
scen_low <- interpolate_model(make_mod(c(50, 50, 50)), name = "BASE",
                              overwrite = TRUE) |>
  solve_scenario(solver = solver_options$glpk, echo = FALSE, variant = "low")
scen_low <- save_scenario(scen_low, verbose = FALSE)
scenario_runs(scen_low)[, c("run", "variant", "modinp", "objective")]
```

```
runs/
├── glpk/  glpk-alt/     # base-problem runs (modinp: shared)
└── low/                 # an own-problem variant
    ├── variant.yml      #   what this problem IS (type, horizon, calendar;
    │                    #   drivers add fields like sequence/step)
    ├── modInp/          #   ITS parameter store — written once, shared by
    ├── variant.RData    #   all of the variant's solves
    └── glpk/…           #   its runs, same shape as any run
```

The base problem's store stays untouched at the scenario level.
`read_solution(scen, run = "glpk")` vs `run = "low/glpk"` switches not just
the solution but the whole in-memory *problem* — settings and parameters
swap together, in both directions — so `getData()` is always consistent
with the run you are looking at. This is the storage substrate for the
sequential (myopic) solver: `solve_myopic()` keeps a whole year-by-year
sequence as one scenario with one variant per step.

## Scenario, variant, or run?

The three axes have crisp definitions — *what changed decides where it
lives*:

| what changed | it is a… | lives at |
|---|---|---|
| solver, tolerances, method, threads | **run** | `runs/[<variant>/]<label>/` |
| the *scope* of the same declarations — a sampled calendar, a region subset or aggregation, a horizon window, a myopic step | **variant** (own problem) | `runs/<variant>/` with its own `modInp` |
| *assumptions or declarations* — a fuel price, a policy cap, a technology's cost | **scenario** | its own folder |

Two runs of one problem agree on every number going in; anything that
changes the calendar or geoscale changes the compiled LP and is therefore
a different *problem*, never a different run; and an assumption change
deserves its own scenario — that is the level `compare_scenarios()`
compares, with its own registry identity. Run and variant names are
**labels** (dashes allowed: `glpk-alt`); scenario/model/dataset names are
**object names** and follow the name rule (letters, digits, underscore).

# A worked example: a UTOPIA project

A project with two policy scenarios (`BASE`, `CO2CAP`) on the all-regions
model `UTOPIA_R11`, solved at full and sampled time resolution, for region
subsets, and for a one-region continental aggregation:

```
utopia-project/
├── energyRt_registry.csv                 the index (rebuildable)
├── datasets/
│   ├── utopia_weather/  utopia_demand/   big tables — stored ONCE
│   └── utopia_map/                       the Geoscale (rds)
├── repositories/
│   └── utopia_r11/                       kit repo; tables as dataset refs
├── models/
│   └── UTOPIA_R11/                       ONE folder, updated in place
├── scenarios/
│   ├── BASE-UTOPIA_R11-d365_h24/         {scenario}-{model}-{calendar}
│   │   ├── scenario.yml · scen.RData     manifest · identity + BASE problem
│   │   ├── modInp/                       base problem's parameters
│   │   └── runs/
│   │       ├── julia_highs/              base-problem solve, full year
│   │       ├── s4_h24/                   VARIANT: sampled calendar
│   │       │   ├── variant.yml · variant.RData · modInp/
│   │       │   ├── glpk/  glpk-barrier/  its solves
│   │       ├── r3/   r1/                 VARIANTS: region subsets
│   │       │   └── … · glpk/
│   │       └── c1/                       VARIANT: continental aggregation
│   │           └── … · julia_highs/
│   ├── CO2CAP-UTOPIA_R11-d365_h24/       the policy scenario — only the
│   │   └── runs/julia_highs/  s4_h24/…   cuts actually solved
│   └── CHEAPGAS-UTOPIA_R11-d365_h24/     an ASSUMPTION change = scenario
└── reports/  levcosts/                   derived-artifact tiers
```

Reading the tree with the definitions above: `s4_h24`, `r3`, `r1` and `c1`
are variants *because* their calendar or geoscale differs; `glpk` and
`glpk-barrier` are runs *because* only the solver differs; `CHEAPGAS` is a
scenario *because* an input price changed. The continental aggregation is
a problem *of `UTOPIA_R11`* — aggregation happens at interpolation, and
`variant.yml` records the coarse geoscale as provenance (only an
explicitly built and stored aggregate model would earn its own `models/`
entry). The weather year exists exactly once, in `datasets/`, referenced
up the chain.

The registry indexes it all — one plain row per thing, run rows carrying
the solve label in `name` and the problem in `variant`:

| type | name | parent | variant | path |
|---|---|---|---|---|
| dataset | `utopia_weather` | | | `datasets/utopia_weather` |
| repository | `utopia_r11` | | | `repositories/utopia_r11` |
| model | `UTOPIA_R11` | | | `models/UTOPIA_R11` |
| scenario | `BASE` | | | `scenarios/BASE-UTOPIA_R11-d365_h24` |
| run | `julia_highs` | `BASE` | | `…/runs/julia_highs` |
| run | `glpk` | `BASE` | `s4_h24` | `…/runs/s4_h24/glpk` |
| run | `glpk-barrier` | `BASE` | `s4_h24` | `…/runs/s4_h24/glpk-barrier` |
| run | `glpk` | `BASE` | `r3` | `…/runs/r3/glpk` |
| run | `julia_highs` | `BASE` | `c1` | `…/runs/c1/julia_highs` |
| scenario | `CO2CAP` | | | `scenarios/CO2CAP-UTOPIA_R11-d365_h24` |
| run | `julia_highs` | `CO2CAP` | | `…/runs/julia_highs` |

Typical queries against it:

```{r utopia-queries, eval = FALSE}
reg <- load_registry()
find_in_registry(reg, type = "run", parent = "BASE", variant = "s4_h24")
find_in_registry(reg, type = "run", variant = "")     # all base-problem runs
getData(c("BASE", "CO2CAP", "CHEAPGAS"),           # compare by NAME alone
        name = "vEmsFuelTot", comm = "CO2", merge = TRUE)
read_solution(getScenario("BASE"), run = "s4_h24/glpk")  # activate a variant run
```

# Moving folders, older layouts

Because run and store locations are derived and manifests travel with their
folders, the whole project directory can be moved, renamed, or synced
between machines. `load_scenario()` re-bases all paths onto wherever the
folder actually is — and if data genuinely is not where the object expects
it, the failure is an *error naming the fix*, never silently empty results.

Folders written by older energyRt versions (the pre-`runs/` layout with
`script/<solver>/` directories) still load, and migrate in place with one
idempotent call:

```r
upgrade_scenario_layout("scenarios/OLD")
```

which moves each solver directory into `runs/`, backfills the provenance
records from what is on disk, and writes the current manifests.

# Lifecycle: sealing and the deletion queue

Because entries update in place, finished work can be locked:
`seal_model("SM")` (and `seal_repository()`, `seal_dataset()`,
`seal_scenario()`) freezes the entry — identical re-saves stay silent
no-ops, changed content errors until the matching `unseal_*()`. A sealed
scenario is an archive: it loads, `getData()`s and reports freely, but
refuses re-saves, new recorded solves, and `drop_scenario_run()`. Seal the
inputs of anything you publish — a sealed model can never drift out from
under the scenarios that reference it.

The other end of the lifecycle is a soft-delete queue:
`mark_delete(x, importance = )` tags an entry (higher importance = more
valuable = deleted later), `unmark_delete()` clears the tag, and
`delete_marked()` — **dry-run by default** — removes marked entries up to
an importance threshold. Sealed entries are never deleted, even when
marked.

# The toolbox

The storage API by task — every verb resolves registered NAMES, never
paths:

```{r toolbox, eval = FALSE}
## anchor a session on a project (registry + all store roots + tiers)
open_project("path/to/project")
get_scenarios_path(); get_models_path(); get_repositories_path()
get_datasets_path(); get_reports_path(); get_levcost_cache_path()

## save — identical content is a no-op; changed content updates in place
save_model(mod);  save_repository(repo);  save_dataset(x, "wsol")
save_scenario(scen)                    # embed_model / embed_repos /
                                       #   embed_datasets = NULL|TRUE|FALSE
save_dataset(name = "kit", fun = "energyRt::utopia_profiles",
             args = list(regions = "R1"))   # store the CALL + a snapshot

## load (explicit IO) and get* (name-first, scenario shells cached in .scen)
load_model("UTOPIA_R11");  load_scenario("BASE");  load_scenarios()
getScenario("BASE");  getModel("UTOPIA_R11");  getDataset("wsol")
getData(c("BASE", "CO2CAP"), name = "vTechOut", merge = TRUE)

## runs and problems
scenario_runs(scen)                    # incl. variant, peak_mb, updated
scenario_run_info(scen, "glpk")        # one run's full run.yml
read_solution(scen, run = "s4_h24/glpk")   # activate problem + solution
drop_scenario_run(scen, "glpk-alt")
solve_scenario(scen, variant = "low")  # solve as an own-problem variant

## the registry
reg <- load_registry()
find_in_registry(reg, type = "run", parent = "BASE", variant = "s4_h24")
refresh_registry()                     # rebuild the index from manifests

## lifecycle
seal_scenario("BASE");  unseal_scenario("BASE")     # freeze / reopen
mark_delete("OLD", type = "scenario", importance = 1)
delete_marked()                        # DRY RUN by default
delete_marked(max_importance = 1, dry_run = FALSE)

## bookkeeping
set_log_file("energyRt_log.csv");  read_log()       # operation log
upgrade_scenario_layout("scenarios/OLD")            # migrate old folders

## custom folder names (session-local hooks; see ?path_builders)
set_path_builder(
  scenario_dir = \(name, model, calendar, horizon)  # scenario folder slug
    paste(name, calendar, sep = "."),
  store_entry  = \(type, name) paste0(type, "_", name),  # store basenames
  run_label    = \(solver)                          # default solve label
    if (is.null(solver$name)) "run" else solver$name
)
set_path_builder(scenario_dir = FALSE)              # FALSE removes one hook
```

# Summary of the vocabulary

| thing | where | identity | manifest |
|---|---|---|---|
| repository | `repositories/<name>/` | name (hash verifies) | `repository.yml` |
| model | `models/<name>/` | name (hash verifies) | `model.yml` |
| dataset | `datasets/<name>/` | name (hash verifies) | `dataset.yml` |
| scenario | `scenarios/<slug>/` | name | `scenario.yml` |
| variant (own problem) | `<scenario>/runs/<variant>/` | name within scenario | `variant.yml` |
| run (solve) | `<scenario>/runs/[<variant>/]<solve>/` | run id `"<solve>"` / `"<variant>/<solve>"` | `run.yml` |
| registry | `energyRt_registry.csv` | — (rebuildable index) | — |

```{r cleanup, include = FALSE}
unlink(proj, recursive = TRUE)
```
