---
title: "Scenario management: how models, scenarios, and solutions are organized"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE
)
```

An energy-modeling project quickly accumulates artifacts: models built from
shared repositories, scenarios interpolated from them at different scopes,
and solutions produced by different solvers. energyRt organizes all of it
in one project folder with three principles:

1. **One folder per thing, one manifest per folder.** Every saved model,
   repository, scenario, and solver run is a directory with a small YAML
   manifest describing it. The manifests — not any central database — are
   the source of truth; the project registry is a rebuildable index over
   them.
2. **Share by content, not by copy.** Models and repositories are stored
   *content-addressed*: their folder name carries a hash of their content,
   identical content is stored once, and other objects reference it by
   name + hash. Embedding a private copy remains available where
   portability matters — "interlinking to avoid duplication is an option,
   not a rule".
3. **Derive paths, don't trust stored ones.** Where a run's files live is
   always computed from the scenario's location and the run's identity.
   Loading a scenario re-bases every stored path onto the folder it is
   actually loaded from, so moving or renaming folders — or opening a
   project from a different working directory — just works.

This article walks the whole layout bottom-up. The *Workflow* article shows
the same machinery in day-to-day use; this one explains how the pieces fit.

```{r setup}
library(energyRt)
```

Everything lands under a throwaway project folder here; in a real project
these are relative paths at the project root (the defaults are
`scenarios/`, `models/`, `repositories/`, and `energyRt_registry.csv`).

```{r paths}
proj <- file.path(tempdir(), "sm-demo")
set_scenarios_path(file.path(proj, "scenarios"))
set_models_path(file.path(proj, "models"))
set_repositories_path(file.path(proj, "repositories"))
set_registry_file(file.path(proj, "energyRt_registry.csv"))
```

A small model to work with — one region, one technology, three years:

```{r model}
cal <- newCalendar(
  timetable = make_timetable(
    struct = list(ANNUAL = "ANNUAL", HOUR = sprintf("t%02d", 1:4))),
  name = "smc")

make_mod <- function(stock) {
  TECH <- newTechnology(
    "TECH", output = data.frame(comm = "ELC", unit = "GWh"), cap2act = 1,
    capacity = data.frame(region = "R1", year = 2020:2022, stock = stock),
    fixom = data.frame(fixom = 1),
    vintage = data.frame(olife = 50L), optimizeRetirement = FALSE)
  newModel(
    name = "SM", region = "R1", discount = 0, calendar = cal,
    horizon = newHorizon(2020:2022, intervals = c(1, 1, 1)),
    repo = newRepository(
      "smr", newCommodity("ELC", timeframe = "HOUR"), TECH,
      newSupply("BACK", commodity = "ELC", supply = data.frame(cost = 1000)),
      newDemand("DEM", commodity = "ELC", demand = data.frame(demand = 1))))
}
mod <- make_mod(stock = c(100, 100, 100))
```

# The scenario folder

A **scenario** is one interpolated instance of a model: the model compiled
onto a concrete calendar, horizon, and region scope. Solving it — possibly
several times, with several solvers — produces **runs**. On disk:

```
scenarios/<name_model_calendar_horizon>/
├── scenario.yml        # manifest: model reference, calendar, horizon,
│                       #   `default:` — the default run (edit it freely)
├── scen.RData          # the thin S4 shell (all big data lives beside it)
├── class · format · layout · logfile.csv
├── modInp/parameters/<param>/           # the (base) interpolated problem
├── problem.RData       # swap data for the base problem (see Variants)
├── model/…             # the model's data — only when embedded (see below)
└── runs/               # one folder per solver run
    ├── glpk/                            # a base-problem run
    │   ├── run.yml     #   provenance: solver, cmdline, started/finished,
    │   │               #   duration, status, objective, host, user
    │   ├── solver/     #   the solver working dir; the solver writes its
    │   │               #   own solver/output/ — never restructured
    │   └── modOut/variables/<var>/      #   the parsed solution store
    └── julia_highs/…                    # another backend, side by side
```

Three ideas to notice:

* **A run owns its artifacts.** Solving the same scenario with a second
  backend does not overwrite anything: each run keeps its inputs, raw
  outputs, provenance, and parsed solution. The in-memory scenario object
  holds ONE active solution at a time; switching is a read, not a re-solve.
* **The run's location is derived.** Nothing stores "where run X lives" —
  it is always `<scenario>/runs/<run-id>/`, computed on demand. That is why
  a moved scenario folder cannot strand its runs.
* **`default:` in scenario.yml is yours.** It names the run a bare
  `read_solution(scen)` resolves to after loading; a save sets it to the
  active run, and you may edit the file to point elsewhere.

Let's produce that tree. `run = "<label>"` names a run explicitly; without
it, the label defaults to the solver directory name (e.g. `"glpk"`):

```{r solve}
scen <- interpolate_model(mod, name = "BASE") |>
  solve_scen(solver = solver_options$glpk, echo = FALSE)
scen <- solve_scen(scen, solver = solver_options$glpk, echo = FALSE,
                   force = TRUE, run = "glpk-alt")
scenario_runs(scen)[, c("run", "status", "solver_name", "objective", "active")]
```

`scenario_run_info(scen, "glpk")` returns one run's full `run.yml`;
`read_solution(scen, run = "glpk")` switches the active solution;
`scenario_drop_run(scen, "glpk-alt")` deletes a run (the active one only
with `force = TRUE`). An *external* working directory — outside the
scenario, no run record — is available via
`solve_scen(solver.dir = "...")`, and `transient = TRUE` solves in a
throwaway folder.

```{r save}
scen <- save_scenario(scen, verbose = FALSE)
sort(list.files(scen@path))
```

# Models and repositories: stored once, referenced by hash

The model is part of the scenario, and by default a saved scenario embeds
its own copy — self-contained, but wasteful when fifty scenarios share one
model. The **model store** removes the duplication:

```{r model-store}
mod <- save_model(mod, verbose = FALSE)      # models/SM@<hash8>/
scen <- save_scenario(scen, verbose = FALSE) # now saved with a REFERENCE
yaml::read_yaml(file.path(scen@path, "scenario.yml"))$model
```

`model_hash()` identifies a model by its *content*: the hash is computed
with all volatile bookkeeping stripped (internal `@misc` metadata, on-disk
markers, session-specific attributes), so two identically-built models hash
identically and any modeled change produces a new store entry —
`models/SM@a1b2c3d4/` and `models/SM@e5f6a7b8/` can coexist, and a scenario
pins exactly the version it was built from. Saving identical content twice
is a no-op. `load_scenario()` resolves the reference back through the
registry, and errors *loudly* — naming the fix — if the store is gone.
`save_scenario(embed_model = TRUE)` keeps a folder self-contained instead.

One level down, the same pattern applies to **repositories** — the shared
building-block collections that many models have in common:

```{r repo-store}
repo <- mod@data[[1]]
save_repository(repo, verbose = FALSE)        # repositories/smr@<hash8>/
mod2 <- save_model(make_mod(c(100, 100, 100)), overwrite = TRUE,
                   verbose = FALSE)           # repo now saved as a reference
```

`save_model(embed_repos = )` mirrors `embed_model`: with the repository in
its store, the model's entry records a `{name, hash}` reference instead of
a data copy, and `load_model()` re-assembles the chain — a scenario
referencing a model whose repositories are themselves references loads
back fully resolved.

# The registry

Every save also records its object in the project **registry** — one CSV at
the project root indexing models, repositories, scenarios, and runs:

```{r registry}
reg <- registry_load()
reg[, c("type", "name", "parent")]
```

The registry is an *index, never an authority*: `registry_refresh()`
rebuilds it at any time by rescanning the stores and reading the manifests.
`registry_find(reg, type = "model", hash = "a1b2")` looks entries up by
type, name, or (short) hash.

# Variants: several problems in one scenario

A scenario sometimes needs *the same model asked a different question* —
a sampled calendar, a restricted horizon, a myopic step with carried-over
capacity. These are **own-problem variants**: named folders under `runs/`
that carry their *own* interpolated problem, with their solves inside.

```{r variant}
scen_low <- interpolate_model(make_mod(c(50, 50, 50)), name = "BASE",
                              overwrite = TRUE) |>
  solve_scen(solver = solver_options$glpk, echo = FALSE, variant = "low")
scen_low <- save_scenario(scen_low, verbose = FALSE)
scenario_runs(scen_low)[, c("run", "variant", "modinp", "objective")]
```

```
runs/
├── glpk/  glpk-alt/     # base-problem runs (modinp: shared)
└── low/                 # an own-problem variant
    ├── variant.yml      #   what this problem IS (type, horizon, calendar;
    │                    #   drivers add fields like sequence/step)
    ├── modInp/          #   ITS parameter store — written once, shared by
    ├── variant.RData    #   all of the variant's solves
    └── glpk/…           #   its runs, same shape as any run
```

The base problem's store stays untouched at the scenario level.
`read_solution(scen, run = "glpk")` vs `run = "low/glpk"` switches not just
the solution but the whole in-memory *problem* — settings and parameters
swap together, in both directions — so `getData()` is always consistent
with the run you are looking at. This is the storage substrate for the
upcoming sequential (myopic) solver, which keeps a whole year-by-year
sequence as one scenario with one variant per step.

# Moving folders, older layouts

Because run and store locations are derived and manifests travel with their
folders, the whole project directory can be moved, renamed, or synced
between machines. `load_scenario()` re-bases all paths onto wherever the
folder actually is — and if data genuinely is not where the object expects
it, the failure is an *error naming the fix*, never silently empty results.

Folders written by older energyRt versions (the pre-`runs/` layout with
`script/<solver>/` directories) still load, and migrate in place with one
idempotent call:

```r
scenario_upgrade_layout("scenarios/OLD")
```

which moves each solver directory into `runs/`, backfills the provenance
records from what is on disk, and writes the current manifests.

# Summary of the vocabulary

| thing | where | identity | manifest |
|---|---|---|---|
| repository | `repositories/<name>@<hash8>/` | content hash | `repository.yml` |
| model | `models/<name>@<hash8>/` | content hash | `model.yml` |
| scenario | `scenarios/<slug>/` | name | `scenario.yml` |
| variant (own problem) | `<scenario>/runs/<variant>/` | name within scenario | `variant.yml` |
| run (solve) | `<scenario>/runs/[<variant>/]<solve>/` | run id `"<solve>"` / `"<variant>/<solve>"` | `run.yml` |
| registry | `energyRt_registry.csv` | — (rebuildable index) | — |

```{r cleanup, include = FALSE}
unlink(proj, recursive = TRUE)
```
