---
title: "Space resolution: geoscales and geoframes"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  message = FALSE, warning = FALSE
)
```

```{r setup}
library(energyRt)
library(geoscales)
```

# Why space has a resolution too

A model's `region` vector is a flat list of labels, and every commodity is
balanced in every one of them. That is the right default for electricity: a
megawatt-hour in one state is not a megawatt-hour in the next, and moving it
needs a line with a capacity and a loss.

It is the wrong default for a lot of other things. Steel, cement, hydrogen
certificates, a national CO~2~ allowance, a country-wide gas market — these are
traded at a scale where the transport inside the scale is cheap, uncongested and
uninteresting. Modelling them per state forces you to invent a transport network
you do not believe in, and to calibrate it.

`commodity@geoframe` says so directly: **balance this commodity at a coarser
level than the model's regions.** It is the spatial twin of
`commodity@timeframe`, which does the same thing in time (see the
[time resolution article](time-resolution.html)).

# The geoscale: what the levels are

A `Geoscale` describes how the model's regions nest. It is built from a
**leaftable** — one row per finest region, one column per level — plus an
ordered `geoframes` vector, coarsest first:

```{r geoscale}
gs <- geoscale_from_leaftable(
  data.frame(
    nation = c("NA1", "NA1", "NA2", "NA2"),
    region = c("R1",  "R2",  "R3",  "R4")
  ),
  geoframes = c("nation", "region"),
  key = "region",
  name = "demo"
)

geoscale_geoframes(gs)
#> [1] "nation" "region"
```

The finest geoframe holds the model's own regions; everything coarser is a
grouping of it. Attach it to the model:

```{r attach}
mod <- newModel("demo", region = c("R1", "R2", "R3", "R4"), geoscale = gs)
# or, on an existing model:
mod <- setGeoscale(mod, gs)
```

Attaching a geoscale **changes nothing on its own**. It widens the model's
region set so `NA1`/`NA2` are members alongside `R1`–`R4`, and those extra
members stay completely inert until some commodity names a coarser level.

# Declaring a commodity at a coarser level

```{r declare}
ELC <- newCommodity("ELC", timeframe = "SEASON")                      # per region
STL <- newCommodity("STL", timeframe = "ANNUAL", geoframe = "nation")  # per nation
```

`ELC` is balanced in R1, R2, R3 and R4 separately, as always. `STL` is balanced
in `NA1` and `NA2` only. A steel mill in R1 and a steel demand in R2 now meet in
the same balance equation, with nothing in between.

## What that actually asserts

**A coarse balance is exactly equivalent to free, unlimited, lossless transport
of that commodity inside the level.** That is not an approximation of the
feature — it is its definition, and the package pins it: the objective of a
model with `STL` at `"nation"` equals, to 1e-9, the objective of the same model
with `STL` per region plus a free unlimited trade route between them.

So the question to ask before using it is never "is this commodity regional?"
but **"would I be willing to draw a costless, uncongested, lossless pipe between
every pair of regions inside this level?"** For an integrated national steel
market, yes. For electricity, no — and the slot documentation says so.

Mechanically: the finer regions' flows are summed into the coarse cell by an
**unweighted** sum. Regional quantities are extensive — they add up — so unlike
the timeslice side there is no weighting parameter. Multi-level chains
accumulate one level at a time, so with `nation > zone > region` the zone cell
must exist for the nation cell to have anything to read; energyRt builds those
intermediate cells for you.

# What you can and cannot do

| | |
|---|---|
| **commodity** at a coarse level | ✅ this is the feature |
| **supply**, **demand**, **storage** at a coarse level | ✅ supported and priced correctly |
| **technology** at a coarse level | ✅ *if* every commodity it touches is at that level or coarser |
| **trade between two coarse regions** | ✅ works — see the FAQ |
| costs (`invcost`, `fixom`) on a coarse cell | ✅ charged once, at that cell |
| **technology** coarser than a commodity it uses | ❌ refused: `"COARSER"` |
| **trade between atoms** for a coarse commodity | ❌ refused: `"Declared resolution does not match"` |
| rest-of-world **import/export** at a coarse level | ❌ refused; an open capability, not a bug |
| **weather** at a coarse level | ❌ outside the level system entirely |
| losses or capacity limits *inside* a level | ❌ impossible by construction — that is what "free" means |
| two **parallel** groupings (trade blocs *and* borders) | ❌ refused; see caveats |

## Strict classes must match exactly

`demand`, `supply`, `import`, `export`, `trade` and `storage` have no
aggregation path of their own. Their flows enter the totals directly, so they
must be declared at **exactly** the commodity's level:

```{r strict}
STL <- newCommodity("STL", timeframe = "ANNUAL", geoframe = "nation")

# Constructs fine, but `interpolate_model()` refuses it:
#   "Declared resolution does not match the commodity's own level"
bad <- newDemand("D", commodity = "STL", region = "R2",
                 demand = data.frame(region = "R2", timeslice = "ANNUAL",
                                     demand = 10))

# Correct: declare it where STL is balanced.
good <- newDemand("D", commodity = "STL", region = "NA1",
                  demand = data.frame(region = "NA1", timeslice = "ANNUAL",
                                      demand = 10))
```

This check exists because the failure it prevents is silent. A demand written at
R2 for a nationally-balanced commodity lands in a cell no equation reads: the
parameter is stored, the balance is generated at the nation, the lookup misses,
and the term evaluates to zero. The model stays feasible and reports no steel
demand at all.

A `technology` is the exception — it may sit **finer** than the commodities it
touches (a mill in one state making national steel), never coarser.

# Pros and cons

**Pros**

- **It removes a network you would otherwise have to invent.** No routes, no
  capacities, no efficiencies, no calibration for transport you do not model.
- **It is smaller.** The same four-region system, modelled as a coarse balance
  inside each nation plus one inter-national link, versus flat regions plus three
  routes: **24 rows / 23 columns / 46 non-zeros** against **52 / 51 / 106**. Less
  than half the LP, for the same answer.
- **It is per commodity.** Steel goes national while electricity stays hourly and
  regional, in one model.
- **Reporting is unaffected.** `getData(geoframe = "nation")` aggregates *any*
  result up a level whether or not the commodity is balanced there.

**Cons**

- **The assumption is strong and invisible in the results.** A coarse balance
  never reports congestion, because there is none. Nothing in the output says
  "this answer assumed free transport".
- **It is all or nothing inside a level.** You cannot make transport inside a
  nation cheap-but-finite. The moment you need a capacity or a loss, you need a
  real `trade` object and a finer balance.
- **It flattens the answer.** Locational detail that would come out of a
  congested network — where the plant gets built, what the regional price is —
  is gone by construction.
- **Tooling is thinner here than for the temporal twin.** Fewer worked examples,
  and no coarse-level test coverage yet for technology, demand or tax/subsidy.

# Use cases

- **A national market inside a sub-national model.** Gas, coal, biomass or steel
  at `"nation"` while electricity stays per region. The common case.
- **A policy instrument with a legal boundary.** A CO~2~ allowance or a
  renewable certificate is issued at a national or continental level; declaring
  the commodity there gives you one cap instead of N caps that do not trade.
- **Continental commodities.** A Europe-wide hydrogen or CO~2~ market in a
  36-country model: with a root geoframe, one market; without it, 36 separate
  ones.
- **Cutting model size in a first pass.** Coarsen the commodities whose
  geography you are not studying yet, and refine later — the declaration is one
  argument, not a rewrite.

# Caveats

**Geoframes must nest.** Adjacent levels must refine one another. `geoscales`
permits cross-cutting frames on purpose, but this roll-up cannot represent them:
the sum is unweighted, so a region with two parents is added into **both**.
energyRt refuses a non-nesting chain and names the offending frames.

**Never repeat a code across adjacent levels.** A region that is its own parent
needs no aggregation, and the pair is dropped with a message. Follow Eurostat
and pad instead (`LU` → `LU0` → `LU00`) if you want the levels visually distinct.

**Do not declare coarse codes in `newModel(region = )`.** They are derived from
the geoscale. Declaring them yourself suppresses the hierarchy.

**Never sum a multi-level result across levels.** Once a geoscale is attached,
totals exist at every level, so `sum(vOutTot$value)` double-counts. Use
`getData(geoframe = )` and take one level.

**A coarse cell inherits its children's discount rate**, and refuses when they
disagree — a deliberate error, not a bug.

# Questions and tips

**Can I avoid declaring `trade` by putting the commodity on an upper level?**

Yes — inside that level, and that is the intended use. Declaring `STL` at
`"nation"` gives you free unlimited movement between `R1` and `R2` with no trade
object at all. What you give up is every property a `trade` carries: capacity,
efficiency (`teff`), cost, and any reporting of the flow. If you need one of
those, you need the route.

**Then how do I move it *between* nations?**

Declare the trade between the coarse regions:

```{r coarse-trade}
newTrade("TR", commodity = "STL",
         routes = data.frame(src = "NA1", dst = "NA2"),
         vintage = data.frame(olife = 50L),
         trade = data.frame(timeslice = "ANNUAL", teff = 1))
```

This is the natural pattern: **free inside the level, an explicit priced link
between levels.** A route between the *atoms* (`R1` → `R3`) is refused for a
nationally-balanced commodity, because the atoms are not where it is balanced.

**Can I put a technology on an upper level?**

Yes, provided every commodity it consumes and produces is balanced at that level
or coarser. A converter at `"nation"` turning national gas into national steel
is fine. The same converter consuming region-level electricity is refused, and
the message names it: the flow could never reach the finer balance.

**Does attaching a geoscale change my results?**

No. It widens the region set and enables coarse *reporting*; the optimisation is
untouched until a commodity names a `@geoframe`. A flat model with a geoscale
attached is bit-for-bit the model without one.

**Can I have one commodity national and another continental?**

Yes. Each commodity names its own level independently, and each gets its own
balance rows. There is no requirement that they agree.

**Can I use two different groupings of the same regions?**

Not today. A model carries one `Geoscale` with one linear chain of geoframes.
Trade blocs alongside administrative borders is a legitimate modelling need and
is refused rather than silently double-counted; keeping the extra grouping as a
leaftable column does not help, because a column that is not in `geoframes` is
invisible to the model.

**How do I check a geoscale before building on it?**

```{r nests}
frames <- geoscale_geoframes(gs)
for (i in seq_len(length(frames) - 1L)) {
  stopifnot(isTRUE(geoscale_nests(gs, frames[i], frames[i + 1L])))
}
```

**How do I see where a commodity is actually balanced?**

After interpolation, `mCommRegion` maps each commodity to the regions it is
balanced at, and `mRegionFamily` holds the parent→child pairs the roll-up uses.
Both are empty in a flat model.

# Choosing a resolution

- Start with everything at the finest level; it is the default and it is never
  wrong, only sometimes bigger than it needs to be.
- Raise a commodity to a coarser level when you can answer yes to the free-pipe
  question, and when you would otherwise be inventing transport data.
- Keep network-constrained carriers — electricity above all — at the finest
  level, and model their transport with `trade`.
- If you find yourself wanting "free, but only up to X", that is a `trade`
  object with a capacity, not a `@geoframe`.
