---
title: "Space resolution: geoscales and geoframes"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  message = FALSE, warning = FALSE
)
```

```{r setup}
library(energyRt)
library(geoscales)
```

# Why space has a resolution too

A model's `region` vector is a flat list of labels, and every commodity is
balanced in every one of them. That is the right default for electricity: a
megawatt-hour in one state is not a megawatt-hour in the next, and moving it
needs a line with a capacity and a loss.

It is the wrong default for a lot of other things. Steel, cement, hydrogen
certificates, a national CO~2~ allowance, a country-wide gas market — these are
traded at a scale where the transport inside the scale is cheap, uncongested and
uninteresting. Modelling them per state forces you to invent a transport network
you do not believe in, and to calibrate it.

`commodity@geoframe` says so directly: **balance this commodity at a coarser
level than the model's regions.** It is the spatial twin of
`commodity@timeframe`, which does the same thing in time (see the
[time resolution article](time-resolution.html)).

# The geoscale: what the levels are

A `Geoscale` describes how the model's regions nest. It is built from a
**leaftable** — one row per finest region, one column per level — plus an
ordered `geoframes` vector, coarsest first:

```{r geoscale}
gs <- geoscale_from_leaftable(
  data.frame(
    nation = c("NA1", "NA1", "NA2", "NA2"),
    region = c("R1",  "R2",  "R3",  "R4")
  ),
  geoframes = c("nation", "region"),
  key = "region",
  name = "demo"
)

geoscale_geoframes(gs)
#> [1] "nation" "region"
```

The finest geoframe holds the model's own regions; everything coarser is a
grouping of it. Attach it to the model:

```{r attach}
mod <- newModel("demo", region = c("R1", "R2", "R3", "R4"), geoscale = gs)
# or, on an existing model:
mod <- setGeoscale(mod, gs)
```

Attaching a geoscale **changes nothing on its own**. It widens the model's
region set so `NA1`/`NA2` are members alongside `R1`–`R4`, and those extra
members stay completely inert until some commodity names a coarser level.

# Declaring a commodity at a coarser level

```{r declare}
ELC <- newCommodity("ELC", timeframe = "SEASON")                      # per region
STL <- newCommodity("STL", timeframe = "ANNUAL", geoframe = "nation")  # per nation
```

`ELC` is balanced in R1, R2, R3 and R4 separately, as always. `STL` is balanced
in `NA1` and `NA2` only. A steel mill in R1 and a steel demand in R2 now meet in
the same balance equation, with nothing in between.

## What that actually asserts

**A coarse balance is exactly equivalent to free, unlimited, lossless transport
of that commodity inside the level.** That is not an approximation of the
feature — it is its definition, and the package pins it: the objective of a
model with `STL` at `"nation"` equals, to 1e-9, the objective of the same model
with `STL` per region plus a free unlimited trade route between them.

So the question to ask before using it is never "is this commodity regional?"
but **"would I be willing to draw a costless, uncongested, lossless pipe between
every pair of regions inside this level?"** For an integrated national steel
market, yes. For electricity, no — and the slot documentation says so.

Mechanically: the finer regions' flows are summed into the coarse cell by an
**unweighted** sum. Regional quantities are extensive — they add up — so unlike
the timeslice side there is no weighting parameter. Multi-level chains
accumulate one level at a time, so with `nation > zone > region` the zone cell
must exist for the nation cell to have anything to read; energyRt builds those
intermediate cells for you.

# What you can and cannot do

Every claim in this section is demonstrated with a model you can run. They share
one setup: four regions in two nations, one year, no discounting, so each
objective is plain arithmetic.

```{r demo-setup}
gs <- geoscale_from_leaftable(
  data.frame(nation = c("NA1", "NA1", "NA2", "NA2"),
             region = c("R1",  "R2",  "R3",  "R4")),
  geoframes = c("nation", "region"), key = "region", name = "demo"
)

cal <- newCalendar(
  timetable = make_timetable(
    struct = list(ANNUAL = "ANNUAL", SEASON = c("WIN", "SUM"))),
  name = "demo_cal"
)

# Build on the four regions, attach the geoscale, solve, return the objective.
solve_demo <- function(name, ...) {
  mod <- newModel(name = name, calendar = cal,
                  region = c("R1", "R2", "R3", "R4"),
                  horizon = newHorizon(2025), discount = 0,
                  repo = newRepository(paste0("repo_", name), ...))
  scen <- interpolate_model(setGeoscale(mod, gs), name = name, overwrite = TRUE)
  sol <- solve_scenario(scen, solver = solver_options$glpk, wait = TRUE)
  sum(getData(sol, "vObjective", merge = TRUE)$value)
}
```

| | |
|---|---|
| **commodity** at a coarse level | ✅ demo 1 |
| **supply**, **demand**, **storage** at a coarse level | ✅ demo 2 |
| **technology** at a coarse level | ✅ if its commodities are at that level or coarser — demo 3 |
| **trade between two coarse regions** | ✅ demo 4 |
| costs (`invcost`, `fixom`) on a coarse cell | ✅ charged once — demo 5 |
| **technology** coarser than a commodity it uses | ❌ refused — demo 3 |
| **trade between atoms** for a coarse commodity | ❌ refused — demo 4 |
| rest-of-world **import** / **export** at a coarse level | ❌ refused — demo 6 |
| **weather** at a coarse region, read by child processes | ✅ one series serves every child — demo 7 |
| losses or capacity limits *inside* a level | ❌ impossible by construction — demo 8 |
| two **parallel** groupings of the same regions | ❌ refused — demo 9 |

## Demo 1 — a commodity at a coarse level, and what it replaces

A mill in R1, steel demanded across the whole of nation NA1, and **no trade
object anywhere**:

```{r d1a}
COA  <- newCommodity("COA", timeframe = "ANNUAL")
ELC  <- newCommodity("ELC", timeframe = "SEASON")
SUP  <- newSupply("SUP", commodity = "COA",
                  supply = data.frame(region = c("R1", "R2", "R3", "R4"),
                                      cost = 1))
ECOA <- newTechnology("ECOA", input = list(comm = "COA"),
                      output = list(comm = "ELC"),
                      invcost = data.frame(invcost = 0),
                      vintage = data.frame(olife = 30L), cap2act = 1)
MILL <- newTechnology("MILL", input = list(comm = "ELC"),
                      output = list(comm = "STL"), region = "R1",
                      ceff = data.frame(comm = "ELC", cinp2use = 1),
                      invcost = data.frame(invcost = 0),
                      vintage = data.frame(olife = 30L), cap2act = 1)

solve_demo("d1a",
  COA, ELC, newCommodity("STL", timeframe = "ANNUAL", geoframe = "nation"),
  SUP, ECOA, MILL,
  newDemand("D", commodity = "STL", region = "NA1",
            demand = data.frame(region = "NA1", timeslice = "ANNUAL",
                                demand = 10)))
#> 10
```

The mill is in R1, the demand is national, and the two meet. The same system
with `STL` left at the finest level and the demand in R2:

```{r d1b}
solve_demo("d1b",
  COA, ELC, newCommodity("STL", timeframe = "ANNUAL"),
  SUP, ECOA, MILL,
  newDemand("D", commodity = "STL", region = "R2",
            demand = data.frame(region = "R2", timeslice = "ANNUAL",
                                demand = 10)))
#> INFEASIBLE -- nothing carries steel from R1 to R2
```

Add a free, unlimited route and it comes back, at the same cost as the coarse
version:

```{r d1c}
solve_demo("d1c",
  COA, ELC, newCommodity("STL", timeframe = "ANNUAL"),
  SUP, ECOA, MILL,
  newTrade("TR", commodity = "STL",
           routes = data.frame(src = "R1", dst = "R2"),
           vintage = data.frame(olife = 50L),
           trade = data.frame(timeslice = "ANNUAL", teff = 1)),
  newDemand("D", commodity = "STL", region = "R2",
            demand = data.frame(region = "R2", timeslice = "ANNUAL",
                                demand = 10)))
#> 10   -- identical to d1a
```

**That triple is the whole feature.** `geoframe = "nation"` buys exactly what a
free unlimited route would have bought, and nothing else.

## Demo 2 — supply, demand and storage at a coarse level

Declared at the level the commodity is balanced at, they are honoured and
priced:

```{r d2a}
solve_demo("d2a",
  newCommodity("GAS", timeframe = "ANNUAL", geoframe = "nation"),
  newSupply("S", commodity = "GAS", region = "NA1",
            supply = data.frame(region = "NA1", cost = 2)),
  newDemand("D", commodity = "GAS", region = "NA1",
            demand = data.frame(region = "NA1", timeslice = "ANNUAL",
                                demand = 10)))
#> 20   -- 10 units x cost 2
```

Storage works there too. Gas is cheap in winter (1) and dear in summer (100), so
it buys 10 in winter and carries 5 across, paying 5 for the out-capacity:

```{r d2b}
solve_demo("d2b",
  newCommodity("GAS", timeframe = "SEASON", geoframe = "nation"),
  newSupply("S", commodity = "GAS", region = "NA1",
            supply = data.frame(region = "NA1", timeslice = c("WIN", "SUM"),
                                cost = c(1, 100))),
  newStorage("STG", commodity = "GAS", region = "NA1", olife = 1L,
             cap2act = 1, invcost = list(out.invcost = 1),
             seff = data.frame(inpeff = 1, outeff = 1)),
  newDemand("D", commodity = "GAS", region = "NA1",
            demand = data.frame(region = "NA1", timeslice = c("WIN", "SUM"),
                                demand = 5)))
#> 15   -- 10 x 1 bought in winter, + out-capacity 5 x 1
```

These are the **strict classes** (`demand`, `supply`, `import`, `export`,
`trade`, `storage`): they have no aggregation path of their own, so they must
name exactly the commodity's level. Writing the demand above at `R2` instead of
`NA1` is refused by `interpolate_model()` with *"Declared resolution does not
match the commodity's own level"*.

The check exists because the failure it prevents is silent: the parameter would
be stored in a cell no equation reads, the balance would be generated at the
nation, the lookup would miss, and the model would solve happily with no steel
demand at all.

## Demo 3 — a technology at a coarse level

Allowed, as long as every commodity it touches is at that level or coarser:

```{r d3a}
solve_demo("d3a",
  newCommodity("GAS", timeframe = "ANNUAL", geoframe = "nation"),
  newCommodity("STL", timeframe = "ANNUAL", geoframe = "nation"),
  newSupply("S", commodity = "GAS", region = "NA1",
            supply = data.frame(region = "NA1", cost = 2)),
  newTechnology("MILL", input = list(comm = "GAS"), output = list(comm = "STL"),
                region = "NA1", invcost = data.frame(invcost = 0),
                vintage = data.frame(olife = 30L), cap2act = 1),
  newDemand("D", commodity = "STL", region = "NA1",
            demand = data.frame(region = "NA1", timeslice = "ANNUAL",
                                demand = 10)))
#> 20
```

Give that same national mill a **region-level** input and it is refused:

```{r d3b}
solve_demo("d3b",
  newCommodity("ELC", timeframe = "SEASON"),
  newCommodity("STL", timeframe = "ANNUAL", geoframe = "nation"),
  newSupply("S", commodity = "ELC",
            supply = data.frame(region = c("R1", "R2"), cost = 2)),
  newTechnology("MILL", input = list(comm = "ELC"), output = list(comm = "STL"),
                region = "NA1", invcost = data.frame(invcost = 0),
                vintage = data.frame(olife = 30L), cap2act = 1),
  newDemand("D", commodity = "STL", region = "NA1",
            demand = data.frame(region = "NA1", timeslice = "ANNUAL",
                                demand = 10)))
#> Error: Process(es) placed at a region level COARSER than a commodity they
#> use, whose flows could never reach that commodity's balance
```

A technology may always sit *finer* than its commodities — a mill in one state
making national steel, as in demo 1 — never coarser.

## Demo 4 — trade between coarse regions, not between atoms

Gas is national, supplied in NA1 and demanded in NA2, so the route joins the two
**nations**:

```{r d4}
gas_between <- function(name, routes) solve_demo(name,
  newCommodity("GAS", timeframe = "ANNUAL", geoframe = "nation"),
  newSupply("S", commodity = "GAS", region = "NA1",
            supply = data.frame(region = "NA1", cost = 2)),
  newDemand("D", commodity = "GAS", region = "NA2",
            demand = data.frame(region = "NA2", timeslice = "ANNUAL",
                                demand = 10)),
  newTrade("TR", commodity = "GAS", routes = routes,
           vintage = data.frame(olife = 50L),
           trade = data.frame(timeslice = "ANNUAL", teff = 1)))

gas_between("d4a", data.frame(src = "NA1", dst = "NA2"))
#> 20

gas_between("d4b", data.frame(src = "R1", dst = "R3"))
#> Error: Declared resolution does not match the commodity's own level
```

**Free inside the level, an explicit priced link between levels.** A route
between atoms is refused, because the atoms are not where this commodity is
balanced.

## Demo 5 — a cost on a coarse cell is charged once

A fixed corridor of 10 units with `fixom = 1`, and supply costing 1. The only
thing that changes is *where* the `fixom` is declared:

```{r d5}
corridor <- function(name, fixom) solve_demo(name,
  newCommodity("ELC", timeframe = "SEASON"),
  newSupply("S", commodity = "ELC",
            supply = data.frame(region = "R1", cost = 1)),
  newTrade("TR", commodity = "ELC", routes = data.frame(src = "R1", dst = "R3"),
           vintage = data.frame(olife = 50L), cap2act = 1,
           capacity = data.frame(cap.fx = 10), fixom = fixom,
           trade = data.frame(timeslice = c("WIN", "SUM"), teff = 1)),
  newDemand("D", commodity = "ELC",
            demand = data.frame(region = "R3", timeslice = c("WIN", "SUM"),
                                demand = 5)))

corridor("d5a", data.frame(region = "R1",  fixom = 1))   # one endpoint
#> 20    -- supply 10, fixom 10

corridor("d5b", data.frame(region = "NA1", fixom = 1))   # a COARSE cell
#> 20    -- charged ONCE at NA1, not once per child region

corridor("d5c", data.frame(fixom = 1))                   # no region named
#> 30    -- broadcasts to BOTH endpoints: fixom 20
```

The third line is the control. When a cost really does land in two places the
objective visibly moves, so the equality of the first two is evidence rather
than coincidence.

## Demo 6 — rest-of-world import and export

```{r d6}
solve_demo("d6",
  newCommodity("GAS", timeframe = "ANNUAL", geoframe = "nation"),
  newImport("IMP", commodity = "GAS", imp = data.frame(price = 2)),
  newDemand("D", commodity = "GAS", region = "NA1",
            demand = data.frame(region = "NA1", timeslice = "ANNUAL",
                                demand = 10)))
#> Error: Declared resolution does not match the commodity's own level
```

`import` and `export` carry no level of their own — their regions resolve to the
atoms — so a coarse commodity is refused. This is a pinned refusal, an open
capability rather than a defect. Where you only need a priced source, use a
`supply` at the coarse level instead, as in demo 2.

## Demo 7 — one weather series shared by a whole level

Weather data is usually the heaviest thing in a model, and a profile that is
the same across a group of regions used to be copied once per region. Declare
it at the parent instead and every child reads the one copy.

A plant in R1 whose summer availability is halved (`wval = 0.5`):

```{r d7}
plant <- function(name, wreg) solve_demo(name,
  newCommodity("ELC", timeframe = "SEASON"),
  newCommodity("COA", timeframe = "ANNUAL"),
  newSupply("S", commodity = "COA",
            supply = data.frame(region = "R1", cost = 0)),
  newWeather("WX", region = wreg, timeframe = "SEASON",
             weather = data.frame(region = wreg, timeslice = c("WIN", "SUM"),
                                  wval = c(1, 0.5))),
  newTechnology("EC", input = list(comm = "COA"), output = list(comm = "ELC"),
                region = "R1", weather = data.frame(weather = "WX", waf.up = 1),
                invcost = data.frame(invcost = 10),
                vintage = data.frame(olife = 30L), cap2act = 1),
  newDemand("D", commodity = "ELC",
            demand = data.frame(region = "R1", timeslice = c("WIN", "SUM"),
                                demand = 5)))

plant("d7a", "R1")    # the weather sits on the atom the plant is in
#> 6.666667   -- vs 3.333333 with no weather: summer wval = 0.5 binds

plant("d7b", "NA1")   # the SAME series, declared once at the nation
#> 6.666667   -- identical; R1 reads the parent's series
```

The lookup goes through `mWeatherRegionAt`, which resolves each process to its
own region when the weather serves it, and otherwise to the **nearest ancestor**
that does. So a nation-level profile feeds all of that nation's regions from a
single stored series, and a flat model resolves through identity rows to exactly
the old behaviour.

The profile has to be *reachable*, though — a weather at `NA2` is no use to a
plant in R1:

```{r d7c}
plant("d7c", "NA2")   # a different nation: nothing to read
#> Error: Weather link(s) that cannot be resolved: 1 (process, region) cell(s)
#> reference a weather object with no series there and no ancestor to read.
```

That is deliberately an error. `pWeather` defaults to 0 and the factor is
multiplicative, so an unresolved link would quietly multiply the availability
bound by zero — the process would be shut down and the model would still solve.

`pWeather` is an **input**, so it is reported at the region it is declared at
(`getData()` returns `NA1` here), while results stay at the processes' own
regions.

## Demo 8 — transport inside a level is free, and only free

The same system three ways: a lossless route, a lossy route, and the coarse
balance.

```{r d8}
# COA, ELC, SUP, ECOA and MILL are the objects built in demo 1.
flat_route <- function(name, teff) solve_demo(name,
  COA, ELC, newCommodity("STL", timeframe = "ANNUAL"), SUP, ECOA, MILL,
  newTrade("TR", commodity = "STL", routes = data.frame(src = "R1", dst = "R2"),
           vintage = data.frame(olife = 50L),
           trade = data.frame(timeslice = "ANNUAL", teff = teff)),
  newDemand("D", commodity = "STL", region = "R2",
            demand = data.frame(region = "R2", timeslice = "ANNUAL",
                                demand = 10)))

flat_route("d8a", 1)     # lossless route
#> 10

flat_route("d8b", 0.8)   # lossy route: make 12.5 to deliver 10
#> 12.5

# and demo 1a -- STL at "nation", no route at all -- was also 10.
```

The coarse balance reproduces the **lossless** case exactly. There is no way to
express the 12.5 — a loss, a capacity or a tariff inside the level — because a
coarse balance has no knob for it. The moment you need one, you need a real
`trade` object and a finer balance.

## Demo 9 — parallel groupings are refused

`zone` and `sync` are two different partitions of the same four regions, so each
sync zone straddles both zones:

```{r d9}
gx <- geoscale_from_leaftable(
  data.frame(zone   = c("W",  "W",  "E",  "E"),
             sync   = c("S1", "S2", "S1", "S2"),
             region = c("R1", "R2", "R3", "R4")),
  geoframes = c("zone", "sync", "region"), key = "region", name = "xcut")

mod <- newModel("x", calendar = cal, region = c("R1", "R2", "R3", "R4"),
                horizon = newHorizon(2025), discount = 0,
                repo = newRepository("repo_x",
                  newCommodity("GAS", timeframe = "ANNUAL")))

interpolate_model(setGeoscale(mod, gx), name = "x")
#> Error: The geoscale's `geoframes` do not nest: 2 region(s) have more than
#> one parent.
#>    S1 <- E + W
#>    S2 <- E + W
#> Adjacent geoframes must refine one another. Parallel groupings of the same
#> regions (e.g. trade blocs alongside national borders) are not supported in
#> one chain: keep a single hierarchy in `geoframes`.
```

The roll-up is an unweighted sum, so a region with two parents would be added
into both. Rather than double-count silently, energyRt refuses.

# Pros and cons

**Pros**

- **It removes a network you would otherwise have to invent.** No routes, no
  capacities, no efficiencies, no calibration for transport you do not model.
- **It is smaller.** The same four-region system, modelled as a coarse balance
  inside each nation plus one inter-national link, versus flat regions plus three
  routes: **24 rows / 23 columns / 46 non-zeros** against **52 / 51 / 106**. Less
  than half the LP, for the same answer.
- **It is per commodity.** Steel goes national while electricity stays hourly and
  regional, in one model.
- **Reporting is unaffected.** `getData(geoframe = "nation")` aggregates *any*
  result up a level whether or not the commodity is balanced there.

**Cons**

- **The assumption is strong and invisible in the results.** A coarse balance
  never reports congestion, because there is none. Nothing in the output says
  "this answer assumed free transport".
- **It is all or nothing inside a level.** You cannot make transport inside a
  nation cheap-but-finite. The moment you need a capacity or a loss, you need a
  real `trade` object and a finer balance.
- **It flattens the answer.** Locational detail that would come out of a
  congested network — where the plant gets built, what the regional price is —
  is gone by construction.
- **Tooling is thinner here than for the temporal twin.** Fewer worked examples,
  and no coarse-level test coverage yet for technology, demand or tax/subsidy.

# Use cases

- **A national market inside a sub-national model.** Gas, coal, biomass or steel
  at `"nation"` while electricity stays per region. The common case.
- **A policy instrument with a legal boundary.** A CO~2~ allowance or a
  renewable certificate is issued at a national or continental level; declaring
  the commodity there gives you one cap instead of N caps that do not trade.
- **Continental commodities.** A Europe-wide hydrogen or CO~2~ market in a
  36-country model: with a root geoframe, one market; without it, 36 separate
  ones.
- **Cutting model size in a first pass.** Coarsen the commodities whose
  geography you are not studying yet, and refine later — the declaration is one
  argument, not a rewrite.

# Caveats

**Geoframes must nest.** Adjacent levels must refine one another. `geoscales`
permits cross-cutting frames on purpose, but this roll-up cannot represent them:
the sum is unweighted, so a region with two parents is added into **both**.
energyRt refuses a non-nesting chain and names the offending frames.

**Never repeat a code across adjacent levels.** A region that is its own parent
needs no aggregation, and the pair is dropped with a message. Follow Eurostat
and pad instead (`LU` → `LU0` → `LU00`) if you want the levels visually distinct.

**Do not declare coarse codes in `newModel(region = )`.** They are derived from
the geoscale. Declaring them yourself suppresses the hierarchy.

**Never sum a multi-level result across levels.** Once a geoscale is attached,
totals exist at every level, so `sum(vOutTot$value)` double-counts. Use
`getData(geoframe = )` and take one level.

**A coarse cell inherits its children's discount rate**, and refuses when they
disagree — a deliberate error, not a bug.

# Questions and tips

**Can I avoid declaring `trade` by putting the commodity on an upper level?**

Yes — inside that level, and that is the intended use. Declaring `STL` at
`"nation"` gives you free unlimited movement between `R1` and `R2` with no trade
object at all. What you give up is every property a `trade` carries: capacity,
efficiency (`teff`), cost, and any reporting of the flow. If you need one of
those, you need the route.

**Then how do I move it *between* nations?**

Declare the trade between the coarse regions:

```{r coarse-trade}
newTrade("TR", commodity = "STL",
         routes = data.frame(src = "NA1", dst = "NA2"),
         vintage = data.frame(olife = 50L),
         trade = data.frame(timeslice = "ANNUAL", teff = 1))
```

This is the natural pattern: **free inside the level, an explicit priced link
between levels.** A route between the *atoms* (`R1` → `R3`) is refused for a
nationally-balanced commodity, because the atoms are not where it is balanced.

**Can I put a technology on an upper level?**

Yes, provided every commodity it consumes and produces is balanced at that level
or coarser. A converter at `"nation"` turning national gas into national steel
is fine. The same converter consuming region-level electricity is refused, and
the message names it: the flow could never reach the finer balance.

**Does attaching a geoscale change my results?**

No. It widens the region set and enables coarse *reporting*; the optimisation is
untouched until a commodity names a `@geoframe`. A flat model with a geoscale
attached is bit-for-bit the model without one.

**Can I have one commodity national and another continental?**

Yes. Each commodity names its own level independently, and each gets its own
balance rows. There is no requirement that they agree.

**Can I use two different groupings of the same regions?**

Not today. A model carries one `Geoscale` with one linear chain of geoframes.
Trade blocs alongside administrative borders is a legitimate modelling need and
is refused rather than silently double-counted; keeping the extra grouping as a
leaftable column does not help, because a column that is not in `geoframes` is
invisible to the model.

**How do I check a geoscale before building on it?**

```{r nests}
frames <- geoscale_geoframes(gs)
for (i in seq_len(length(frames) - 1L)) {
  stopifnot(isTRUE(geoscale_nests(gs, frames[i], frames[i + 1L])))
}
```

**How do I see where a commodity is actually balanced?**

After interpolation, `mCommRegion` maps each commodity to the regions it is
balanced at, and `mRegionFamily` holds the parent→child pairs the roll-up uses.
Both are empty in a flat model.

# Choosing a resolution

- Start with everything at the finest level; it is the default and it is never
  wrong, only sometimes bigger than it needs to be.
- Raise a commodity to a coarser level when you can answer yes to the free-pipe
  question, and when you would otherwise be inventing transport data.
- Keep network-constrained carriers — electricity above all — at the finest
  level, and model their transport with `trade`.
- If you find yourself wanting "free, but only up to X", that is a `trade`
  object with a capacity, not a `@geoframe`.
