---
title: "Time resolution: calendars and slices"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("ggplot2", quietly = TRUE),
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, fig.retina = TRUE
)
```

```{r setup}
library(energyRt)
library(ggplot2)
```

# Why sub-annual time matters

Annual averages hide what makes energy systems hard: the sun sets, wind stalls,
demand peaks in the evening. A model that balances electricity once a year sees
none of it — storage is pointless, solar looks dispatchable, peak capacity is
free. A **calendar** gives the model sub-annual **time slices**, and the slice
count is the main dial between realism and model size:

| calendar | structure | slices |
|---|---|---:|
| `utopia_annual` | one annual slice | 1 |
| `utopia_seasons` | 4 seasons × day/night/peak | 12 |
| `utopia_s4h24` | 4 seasons × 24 hours | 96 |
| `utopia_m12h24` | 12 months × 24 hours | 288 |
| `d365` | 365 days | 365 |

Model variables scale roughly linearly with slices — the 96-slice UTOPIA base
case solves in seconds on GLPK, the 288-slice variant is noticeably heavier.

# The `make_timetable()` grammar

A calendar's structure is a **nested named list**: each element is a *level*
(e.g. `SEASON`, `HOUR`), holding its *slices*. Slice names must be
alphanumeric (they become set elements in the solver files). The simplest form
lists slice names — the year is divided equally:

```{r tt-equal}
tt <- make_timetable(list(
  SEASON = c("WIN", "SPR", "SUM", "AUT"),
  HOUR   = paste0("h", formatC(0:23, width = 2, flag = "0"))
))
head(tt)          # 4 x 24 = 96 leaf slices, equal shares
```

Unequal **shares** are given per slice; a nested `list(<share>, <LEVEL> = ...)`
attaches child levels. UTOPIA's 12-slice calendar makes peak hours short and
winter nights long:

```{r tt-shares}
tt12 <- make_timetable(list(
  SEASON = list(
    WIN = list(1 / 4, HOUR = list(DAY =  9 / 24, NGT = 12 / 24, PK = 3 / 24)),
    SPR = list(1 / 4, HOUR = list(DAY = 11 / 24, NGT = 11 / 24, PK = 2 / 24)),
    SUM = list(1 / 4, HOUR = list(DAY = 12 / 24, NGT =  9 / 24, PK = 3 / 24)),
    AUT = list(1 / 4, HOUR = list(DAY = 11 / 24, NGT = 11 / 24, PK = 2 / 24))
  )
))
head(tt12)
```

# From timetable to calendar

`newCalendar()` turns a timetable into a `calendar` object. Note that `name`
is the *first* argument — always pass the timetable by name (`timetable =`),
or name `name`/`desc` so the timetable lands in the right slot:

```{r calendar}
cal <- newCalendar(timetable = tt, name = "s4h24")
cal@name
nrow(cal@slice_share)          # slices with their share of the year
head(as.data.frame(cal@slice_share), 3)
cal@timeframe_rank             # levels, coarsest (ANNUAL) to finest
```

The derived slots do the bookkeeping the model needs:

* **`@slice_share`** — each slice's share of the year (the weight used
  whenever slice values are aggregated);
* **`@timeframes`** — the slice sets at every level (`ANNUAL`, `SEASON`, ...);
* **`@timeframe_rank`** — the level hierarchy; a commodity's `timeframe`
  picks the level it is balanced on.

`autoplot()` draws the nested structure:

```{r cal-plot, fig.height = 5}
autoplot(cal)
```

# Ready-made calendars

The package ships a `calendars` list, built by `data-raw/calendars.R` with
exactly the grammar above — including the whole UTOPIA family:

```{r calendars-data}
names(calendars)
calendars$utopia_seasons@desc
```

Pick one and pass it to `newModel(calendar = ...)`; the UTOPIA vignettes use
`calendars$utopia_s4h24` throughout.

# Slice-string helpers

Slice names encode time; a few helpers translate between encodings:

```{r helpers}
hour2HOUR(c(0, 13, 23))        # hour of day -> "h00" "h13" "h23"
yday2YDAY(c(1, 365))           # day of year -> "d001" "d365"
head(tsl_formats)              # known slice-name formats (a dataset)
```

`tsl2dtm()` / `dtm2tsl()` convert slice strings to/from date-times — useful
when joining model output with observed hourly data.

# Timeframes: commodities and processes

Every commodity carries a `timeframe` — the calendar level it is **balanced**
on. Fuels are typically `ANNUAL`; electricity `HOUR`:

```{r timeframes, eval = FALSE}
ELC <- newCommodity("ELC", timeframe = "HOUR")    # balanced every slice
COA <- newCommodity("COA", timeframe = "ANNUAL")  # balanced once a year
```

A process operates at the *finest* timeframe among its commodities (a gas
plant producing hourly `ELC` is dispatched hourly), overridable via
`newTechnology(timeframe = ...)` — see the
[model bricks article](model-bricks.html) for details. The practical
consequence: raising the calendar's resolution refines *only* the commodities
and processes whose timeframes follow it — annual bookkeeping stays cheap.

# Choosing a resolution

* Start coarse (`utopia_seasons`-like, ~12 slices) while the model structure
  is in flux — solves are instant.
* Move to hour-within-season (`utopia_s4h24`, 96) once storage, VRE profiles
  or peak pricing matter — intra-day dynamics need real hours.
* Full-year hourly detail (`utopia_m12h24`, 288 or `d365`+hours) is for final
  runs; check tractability with `model_size()` first.

The [UTOPIA vignettes](utopia-build.html) build one model and run it on these
calendars interchangeably — resolution is a configuration choice, not a
rewrite.
