---
title: "Workflow: working with results and scenarios"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 3.4, fig.retina = 2, dpi = 110,
  eval = requireNamespace("ggplot2", quietly = TRUE)
)
```

Once a model is **built** (see *Model bricks* and *UTOPIA I*) and **solved** (see
*Solver backends* and *UTOPIA II*), a handful of utility functions do the rest of
the day-to-day work: pull objects and results out, edit a piece and re-solve,
organize scenarios on disk, and compare runs. This article tours that layer.

```{r setup}
library(energyRt)
library(dplyr)
library(ggplot2)
```

We use the packaged single-region UTOPIA kit as a running example, and keep all
scenario files under a temporary folder.

```{r model}
set_scenarios_path(file.path(tempdir(), "wf"))   # where scenarios are written
set_registry_file(file.path(tempdir(), "wf", "energyRt_registry.csv"))

um  <- utopia$modules$electricity$R1
mod <- newModel("UTOPIA", data = um$repo,
                calendar = utopia$modules$calendars$s4_h24,
                region   = um$regions,
                horizon  = utopia$modules$horizons$base, discount = 0.05)
```

```{r base-solve}
# interpolate + solve on the bundled GLPK; echo = FALSE keeps the log quiet
scen <- interpolate_model(mod, name = "BASE") |>
  solve_scenario(solver = solver_options$glpk, echo = FALSE)
```

# Selecting objects: `getObject()`

`getObject()` returns the *building-block objects* held in a repository, model or
scenario, filtered by **class**, **name**, **description**, **region** or any
object slot. It is the object-level counterpart of `getData()` (which returns
their data). By default it returns a named list keyed by object name; `drop =
TRUE` unwraps a single match into the object itself.

```{r getobject}
repo <- um$repo
names(getObject(repo, class = "technology"))        # all technologies
names(getObject(repo, class = "supply", region = "R1"))
getObject(repo, name = "ECOA", drop = TRUE)@invcost # the ECOA object itself
```

Region matching reads `@region` slots and the `region`/`src`/`dst` columns of any
data.frame slot, so it works for every class; region-agnostic objects (e.g.
commodities) match every region. `getObject()` accepts a `scenario` too, so the
same query works before or after solving.

# Extracting data: `getData()`

`getData()` pulls **parameter** and **variable** data out of a scenario as tidy
data.frames. With `merge = TRUE` it returns one long frame (a named list
otherwise); every frame carries a `scenario` and a `name` column.

```{r getdata}
getData(scen, "vObjective", merge = TRUE)$value      # total system cost, MEUR

gen <- getData(scen, "vTechOut", comm = "ELC", merge = TRUE)
head(gen[, c("scenario", "tech", "region", "year", "timeslice", "value")], 4)
```

The `...` accept set filters, exact (`comm = "ELC"`) or regex (`comm_ = "^EL"`).
`timeframe = c("lowest","highest","all")` controls how timeslice-indexed values are
returned. Rename dimensions or recode values on the way out with `newNames =` /
`newValues =` (or afterwards with `renameSets()` / `revalueSets()`), and discover
which parameters carry a given set with `findData()`:

```{r finddata}
names(findData(scen, setsNames_ = "tech"))           # parameters indexed by 'tech'
```

`find_in_model(mod, "ECOA")` text-searches every object slot for a value — handy
for locating where a name is used.

# Editing an object: `update()`

`update()` edits the slots of a **single** model object (a technology, commodity,
demand, ...). It does **not** operate on a whole model or scenario — you update
the object, put it back into the repository with `add(..., overwrite = TRUE)`, and
re-interpolate/solve.

```{r update}
ECOA <- getObject(repo, name = "ECOA", drop = TRUE)
ECOA <- update(ECOA, invcost = data.frame(invcost = 2500))  # pricier coal capex
repo_hi <- add(repo, ECOA, overwrite = TRUE)                # swap it back in

mod_hi  <- newModel("UTOPIA_HI", data = repo_hi,
                    calendar = utopia$modules$calendars$s4_h24,
                    region = um$regions, horizon = utopia$modules$horizons$base,
                    discount = 0.05)
```

(For an already-interpolated scenario, `update_parameter(scen, param, data)`
writes rows straight into its parameter store.)

# Scenario folders and structure

(The *Scenario management* article explains this layer in depth — stores,
manifests, the registry, runs, and variants; this section is the quick tour.)

The scenarios directory is an option — read it with `get_scenarios_path()`, set it
with `set_scenarios_path()` (default `"scenarios/"`). Each scenario gets its own
folder, named automatically as `name-model-calendar-horizon` (parts joined
by dashes, which valid names cannot contain; duplicate parts
are dropped, so a scenario and model sharing a name appear once):

```{r dirname}
get_scenarios_path()
basename(scen@path)
```

Saving a scenario writes an **Arrow-backed** folder that mirrors the scenario's
structure — a thin `scen.RData` shell, a `scenario.yml` manifest, each large
data slot as a dataset, and one folder per solver run:

```
<scenarios-path>/<name-model-calendar-horizon>/
├── scenario.yml        # manifest: model ref/hash, calendar, horizon,
│                       #   `default:` — the default run (user-changeable)
├── scen.RData          # thinned S4 scenario shell
├── class · format · layout   # "scenario" · "parquet" · "3"
├── logfile.csv         # timestamped save log
├── model/data/<repo>/<object>/<slot>/   # model data (embedded scenarios only)
├── modInp/parameters/<param>/           # interpolated input parameters
└── runs/               # one folder per solve
    ├── glpk/           #   base-problem run, named by its solve label
    │   ├── run.yml     #     provenance: solver, cmdline, times, objective
    │   ├── solver/     #     the solver working dir (inputs + its output/)
    │   └── modOut/variables/<var>/      # the solved variables
    └── julia_highs/    #   another backend's run of the same problem
```

Beyond the per-scenario traces above, an optional *operation log* records
the sequence of a working session: `set_log_file("energyRt_log.csv")` makes
`interpolate_model()`, `solve_scenario()` and `solve_myopic()` append one
line each (operation, object, status, objective, duration), and
`read_log()` returns the sequence as a tibble. It is off by default.

Solving the same scenario with several backends (or option sets — pass
`run = "<label>"` to `solve_scenario()`) keeps every run side by side;
`scenario_runs(scen)` lists them with status and objective, and
`read_solution(scen, run = "<label>")` switches the active solution.
A scenario referencing a model saved with `save_model()` skips the embedded
`model/` copy entirely (see `?save_scenario`, `embed_model`).

A scenario knows whether its data is in RAM or on disk via `isInMemory()`; when
on disk, the slots are empty and read lazily from the folder on demand. Folders
from older energyRt versions are still readable and migrate in place with
`upgrade_scenario_layout(path)`.

# Saving and loading

`save_scenario()` spills the scenario to its folder and returns the (now
on-disk) object; `load_scenario(path, env = NULL)` reads the shell back. Its data
stays on disk until requested — `getData()` reads it lazily, or `obj2mem()`
pulls the whole scenario back into memory.

```{r saveload-run, results = "hide"}
saved <- save_scenario(scen, verbose = FALSE)
ld    <- load_scenario(saved@path, env = NULL, verbose = FALSE)
```

A saved scenario is also reachable by its registered **name** —
`load_scenario("BASE", env = NULL)` resolves it through the project
registry, and the cached accessor `getScenario("BASE")` is the everyday
short form (see the *Scenario management* article):

```{r saveload-name}
identical(getScenario("BASE")@name, ld@name)
```

```{r saveload}
basename(saved@path)                                 # the scenario folder
isInMemory(saved)                                    # FALSE -- data is on disk
getData(ld, "vObjective", merge = TRUE)$value        # lazy read, no full load
```

```{r rehydrate-run, results = "hide"}
ld <- obj2mem(ld)                                    # rehydrate fully into RAM
```

```{r rehydrate}
isInMemory(ld)
```

Every `save_scenario()` / `save_model()` / `save_repository()` also records
its object in the project **registry** — a single CSV (default
`energyRt_registry.csv` at the project root, see `get_registry_file()`)
indexing saved models, repositories, scenarios, and runs. Repositories shared
by several models can be stored once with `save_repository()` and referenced
by each model (`save_model(embed_repos = )`), just as scenarios can reference
a stored model. `load_registry()` reads it, `find_in_registry()` filters it (by type,
name, or model hash), and `refresh_registry()` rebuilds it by rescanning the
`scenarios/` and `models/` stores — the on-disk manifests are the source of
truth, so the registry can always be reconstructed.

# Comparing scenarios

Layer a policy lever onto the base model to get a second scenario, then pass a
**named list of scenarios** to `getData()` — the `scenario` column makes the
comparison a one-liner. (Scenarios already *saved* compare by name alone:
`getData(c("BASE", "CO2CAP"), ...)` loads them from the registry on the fly.)

```{r compare-solve}
scen_cap <- interpolate_model(mod, "CO2CAP", um$CO2_CAP) |>   # add the CO2 cap lever
  solve_scenario(solver = solver_options$glpk, echo = FALSE)
```

```{r compare-emis}
emis <- getData(list(BASE = scen, CO2CAP = scen_cap), "vEmsFuelTot",
                comm = "CO2", merge = TRUE)

emis |>
  group_by(scenario, year) |> summarise(ktCO2 = sum(value), .groups = "drop") |>
  ggplot(aes(factor(year), ktCO2, fill = scenario)) +
  geom_col(position = "dodge") +
  labs(x = "year", title = "CO2 emissions by scenario") + theme_bw()
```

```{r compare-cost}
sapply(list(BASE = scen, CO2CAP = scen_cap),
       function(s) round(getData(s, "vObjective", merge = TRUE)$value[1]))
```

To reason about **model size** rather than results, `model_size()` estimates the
parameter/variable/constraint counts of an interpolated scenario, and `size()`
reports its in-memory footprint:

```{r sizes}
model_size(scen)
size(scen)
```

For a systematic check that a build is correct *and* efficient,
`compare_interp_settings(mod)` and `compare_solve_settings(mod, solvers = ...)`
interpolate (and solve) the model under different storage settings and tabulate
build size, time and — crucially — confirm the objective is invariant across
`fold`/`sparse`/`prune`. These run several builds, so they are best run
interactively:

```{r compare-settings, eval = FALSE}
compare_interp_settings(mod)                          # size/time by setting
compare_solve_settings(mod, solvers = list(solver_options$glpk))
```

# Sampling: solving a subset of the model

Interpolation can solve a *sample* of a model without editing the model
itself — in time (a sampled calendar) and in space (a sampled geoscale).
Both work the same way: build the sample object, pass it to
`interpolate_model()`, and the scenario is declared, validated and solved
on the sample.

## Sampled calendars

A sampled calendar is a **row subset of the full calendar's timetable**
plus an explicit `year_fraction` — the fraction of the year the sample
covers. Timeslice weights annualise the sample inside the solver
(the top slice carries weight `1/year_fraction`), so the objective is an
*estimate* of the full model's — and exactly equals it when the sampled
slices are representative:

```{r sample-calendar}
tt  <- utopia$modules$calendars$s4_h24@timetable
tt2 <- tt[tt$SEASON %in% c("WIN", "SUM"), ]      # two of the four seasons
cal2 <- newCalendar(timetable = tt2, name = "s2h24",
                    year_fraction = sum(tt2$share))

scen_t <- interpolate_model(mod, "S2", cal2) |>
  solve_scenario(solver = solver_options$glpk, echo = FALSE)
sapply(list(full = scen, sampled = scen_t),
       function(s) round(getData(s, "vObjective", merge = TRUE)$value[1]))
```

Two traps: pass `year_fraction` **explicitly** (it is not recomputed from
the subset), and **subset the existing timetable** rather than rebuilding
one with `make_timetable()` (which would renormalise the shares to a full
year). Declarations are validated against the *model's* calendar, so
objects naming out-of-sample timeslices interpolate cleanly — their
out-of-sample rows are simply filtered.

## Spatial sampling

The spatial mirror: pass a **filtered geoscale** — a
`geoscales::filter_geoscale()` subset of the model's regions — and the
scenario becomes a *sub-territory* model: only the sampled regions are
declared, out-of-sample parameter rows are filtered, and trade routes
crossing the sample boundary are dropped (with a message). Unlike
calendar sampling there is **no reweighting**: regional quantities are
extensive, so the objective is the sub-territory's own, and disjoint
samples *add up* to the full model (absent cross-boundary trade).

UTOPIA is single-region, so a compact three-region demo:

```{r sample-regions-model}
library(geoscales)

r3 <- c("R1", "R2", "R3")
gs <- geoscale_from_leaftable(
  data.frame(zone = c("Z12", "Z12", "Z3"), region = r3, km2 = 1),
  geoframes = c("zone", "region"), name = "demo")

m3 <- newModel(
  "M3", region = r3, horizon = newHorizon(2020), discount = 0.05,
  repo = newRepository(
    "r",
    newCommodity("COA", timeframe = "ANNUAL"),
    newCommodity("ELC", timeframe = "ANNUAL"),
    newSupply("SUP_COA", commodity = "COA",
              supply = data.frame(region = r3, cost = 5)),
    newTechnology("ECOA", input = list(comm = "COA"),
                  output = list(comm = "ELC"),
                  af = data.frame(af.up = 0.9),
                  invcost = data.frame(invcost = 1000),
                  olife = list(olife = 30), cap2act = 1),
    newDemand("DEM_ELC", commodity = "ELC",
              demand = data.frame(region = r3, demand = c(10, 20, 30))),
    newTrade("TR23", commodity = "ELC", cap2act = 1,
             routes = data.frame(src = "R2", dst = "R3"),
             trade  = data.frame(src = "R2", dst = "R3", ava.up = 4),
             invcost = data.frame(invcost = 10))))
```

Filtering the geoscale to `R1 + R2` and interpolating drops the
`R2 -> R3` route on the way:

```{r sample-regions, message = TRUE}
gs12 <- filter_geoscale(gs, "region", c("R1", "R2"),
                        drop_empty_geoframes = TRUE)
scen12 <- interpolate_model(m3, "S12", gs12)
as.character(scen12@settings@region)
```

A dropped route can optionally leave a **priced boundary stub** at the
kept endpoint — an export (kept source) or import (kept destination)
object bounded by `cap.up` (or, failing that, the route's `ava.up`) —
via `boundary_prices`; routes without a price row are just dropped:

```{r sample-stub, message = TRUE}
bp <- data.frame(src = "R2", dst = "R3", price = 50, cap.up = 4)
scen12p <- interpolate_model(m3, "S12P", gs12, boundary_prices = bp)
grep("^EXP", as.character(scen12p@modInp@sets$expp), value = TRUE)
```

The same surgery is available up front, as a model-to-model verb:

```{r subset-verb}
m12 <- subset_model_regions(m3, c("R1", "R2"), boundary_prices = bp)
as.character(m12@config@region)
```

The filtered geoscale records its **coverage** (fraction of each weight's
parent total — see `geoscales::geoscale_coverage()`), and its name is
mangled (`"demo[region:R1+R2]"`) so different samples of one parent never
collide in caches or joins. A **pruned** geoscale (a coarser level of the
model's own hierarchy) requests a full-territory solve at the parent
level instead; that requires aggregating the model's data
(`aggregate_model_regions()`) and is not implemented yet.

# See also

* **Model bricks** and **UTOPIA I** — building the objects and the model.
* **Solver backends** and **UTOPIA II** — interpolating and solving.
* `?getData`, `?getObject`, `?update`, `?save_scenario`, `?load_scenario`,
  `?model_size`, `?compare_solve_settings`, `?subset_model_regions`.
