---
title: "Workflow: working with results and scenarios"
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 3.4, fig.retina = 2, dpi = 110,
  eval = requireNamespace("ggplot2", quietly = TRUE)
)
```

Once a model is **built** (see *Model bricks* and *UTOPIA I*) and **solved** (see
*Solver backends* and *UTOPIA II*), a handful of utility functions do the rest of
the day-to-day work: pull objects and results out, edit a piece and re-solve,
organize scenarios on disk, and compare runs. This article tours that layer.

```{r setup}
library(energyRt)
library(dplyr)
library(ggplot2)
```

We use the packaged single-region UTOPIA kit as a running example, and keep all
scenario files under a temporary folder.

```{r model}
set_scenarios_path(file.path(tempdir(), "wf"))   # where scenarios are written

um  <- utopia_modules$electricity$reg1
mod <- newModel("UTOPIA", data = um$repo,
                calendar = utopia_modules$calendars$utopia_s4h24,
                region   = um$regions,
                horizon  = utopia_modules$horizons$base, discount = 0.05)
```

```{r base-solve}
# interpolate + solve on the bundled GLPK; echo = FALSE keeps the log quiet
scen <- interpolate_model(mod, name = "BASE") |>
  solve_scenario(solver = solver_options$glpk, echo = FALSE)
```

# Selecting objects: `getObject()`

`getObject()` returns the *building-block objects* held in a repository, model or
scenario, filtered by **class**, **name**, **description**, **region** or any
object slot. It is the object-level counterpart of `getData()` (which returns
their data). By default it returns a named list keyed by object name; `drop =
TRUE` unwraps a single match into the object itself.

```{r getobject}
repo <- um$repo
names(getObject(repo, class = "technology"))        # all technologies
names(getObject(repo, class = "supply", region = "R1"))
getObject(repo, name = "ECOA", drop = TRUE)@invcost # the ECOA object itself
```

Region matching reads `@region` slots and the `region`/`src`/`dst` columns of any
data.frame slot, so it works for every class; region-agnostic objects (e.g.
commodities) match every region. `getObject()` accepts a `scenario` too, so the
same query works before or after solving.

# Extracting data: `getData()`

`getData()` pulls **parameter** and **variable** data out of a scenario as tidy
data.frames. With `merge = TRUE` it returns one long frame (a named list
otherwise); every frame carries a `scenario` and a `name` column.

```{r getdata}
getData(scen, "vObjective", merge = TRUE)$value      # total system cost, MEUR

gen <- getData(scen, "vTechOut", comm = "ELC", merge = TRUE)
head(gen[, c("scenario", "tech", "region", "year", "slice", "value")], 4)
```

The `...` accept set filters, exact (`comm = "ELC"`) or regex (`comm_ = "^EL"`).
`timeframe = c("lowest","highest","all")` controls how slice-indexed values are
returned. Rename dimensions or recode values on the way out with `newNames =` /
`newValues =` (or afterwards with `renameSets()` / `revalueSets()`), and discover
which parameters carry a given set with `findData()`:

```{r finddata}
names(findData(scen, setsNames_ = "tech"))           # parameters indexed by 'tech'
```

`find_in_model(mod, "ECOA")` text-searches every object slot for a value — handy
for locating where a name is used.

# Editing an object: `update()`

`update()` edits the slots of a **single** model object (a technology, commodity,
demand, ...). It does **not** operate on a whole model or scenario — you update
the object, put it back into the repository with `add(..., overwrite = TRUE)`, and
re-interpolate/solve.

```{r update}
ECOA <- getObject(repo, name = "ECOA", drop = TRUE)
ECOA <- update(ECOA, invcost = data.frame(invcost = 2500))  # pricier coal capex
repo_hi <- add(repo, ECOA, overwrite = TRUE)                # swap it back in

mod_hi  <- newModel("UTOPIA_HI", data = repo_hi,
                    calendar = utopia_modules$calendars$utopia_s4h24,
                    region = um$regions, horizon = utopia_modules$horizons$base,
                    discount = 0.05)
```

(For an already-interpolated scenario, `update_parameter(scen, param, data)`
writes rows straight into its parameter store.)

# Scenario folders and structure

The scenarios directory is an option — read it with `get_scenarios_path()`, set it
with `set_scenarios_path()` (default `"scenarios/"`). Each scenario gets its own
folder, named by `make_scenario_dirname()` as `name_model_calendar_horizon`:

```{r dirname}
get_scenarios_path()
make_scenario_dirname(scen)
```

Saving a scenario writes an **Arrow-backed** folder that mirrors the scenario's
structure — a thin `scen.RData` shell plus each large data slot as a dataset:

```
<scenarios-path>/<name_model_calendar_horizon>/
├── scen.RData          # thinned S4 scenario shell
├── class · format      # "scenario" · "parquet"
├── logfile.csv         # timestamped save log
├── model/data/<repo>/<object>/<slot>/   # the model objects' data
├── modInp/parameters/<param>/           # interpolated input parameters
├── modOut/variables/<var>/              # solved variables (vTechOut, ...)
└── script/<backend>_<solver>_<method>/  # solver working directory
```

A scenario knows whether its data is in RAM or on disk via `isInMemory()`; when
on disk, the slots are empty and read lazily from the folder on demand.

# Saving and loading

`save_scenario()` spills the scenario to its folder and returns the (now
on-disk) object; `load_scenario(path, env = NULL)` reads the shell back. Its data
stays on disk until requested — `getData()` reads it lazily, or `obj2mem()`
pulls the whole scenario back into memory.

```{r saveload-run, results = "hide"}
saved <- save_scenario(scen, verbose = FALSE)
ld    <- load_scenario(saved@path, env = NULL, verbose = FALSE)
```

```{r saveload}
basename(saved@path)                                 # the scenario folder
isInMemory(saved)                                    # FALSE -- data is on disk
getData(ld, "vObjective", merge = TRUE)$value        # lazy read, no full load
```

```{r rehydrate-run, results = "hide"}
ld <- obj2mem(ld)                                    # rehydrate fully into RAM
```

```{r rehydrate}
isInMemory(ld)
```

> A name-based scenario **registry** (`register()`, `getScenario()`,
> `get_registry()`) is under development; for now, organize runs with
> `set_scenarios_path()` and round-trip them with `save_scenario()` /
> `load_scenario()`.

# Comparing scenarios

Layer a policy lever onto the base model to get a second scenario, then pass a
**named list of scenarios** to `getData()` — the `scenario` column makes the
comparison a one-liner:

```{r compare-solve}
scen_cap <- interpolate_model(mod, "CO2CAP", um$CO2_CAP) |>   # add the CO2 cap lever
  solve_scenario(solver = solver_options$glpk, echo = FALSE)
```

```{r compare-emis}
emis <- getData(list(BASE = scen, CO2CAP = scen_cap), "vEmsFuelTot",
                comm = "CO2", merge = TRUE)

emis |>
  group_by(scenario, year) |> summarise(ktCO2 = sum(value), .groups = "drop") |>
  ggplot(aes(factor(year), ktCO2, fill = scenario)) +
  geom_col(position = "dodge") +
  labs(x = "year", title = "CO2 emissions by scenario") + theme_bw()
```

```{r compare-cost}
sapply(list(BASE = scen, CO2CAP = scen_cap),
       function(s) round(getData(s, "vObjective", merge = TRUE)$value[1]))
```

To reason about **model size** rather than results, `model_size()` estimates the
parameter/variable/constraint counts of an interpolated scenario, and `size()`
reports its in-memory footprint:

```{r sizes}
model_size(scen)
size(scen)
```

For a systematic check that a build is correct *and* efficient,
`compare_interp_settings(mod)` and `compare_solve_settings(mod, solvers = ...)`
interpolate (and solve) the model under different storage settings and tabulate
build size, time and — crucially — confirm the objective is invariant across
`fold`/`sparse`/`prune`. These run several builds, so they are best run
interactively:

```{r compare-settings, eval = FALSE}
compare_interp_settings(mod)                          # size/time by setting
compare_solve_settings(mod, solvers = list(solver_options$glpk))
```

# See also

* **Model bricks** and **UTOPIA I** — building the objects and the model.
* **Solver backends** and **UTOPIA II** — interpolating and solving.
* `?getData`, `?getObject`, `?update`, `?save_scenario`, `?load_scenario`,
  `?model_size`, `?compare_solve_settings`.
